---
title: "Analysis of the 2020 Presidential Election"
author: "Alexander Sanchez"
date: "2030-01-01"
output: github_document
---


```{r setup, echo=FALSE, include=FALSE}
# install.packages("arulesViz")
library(knitr)
library(ISLR) 
library(tree)
library(maptree)
library(randomForest) 
library(gbm) 
library(glmnet) 
library(caret)
library(psych)
library(boot)
library(tidyverse)
library(corrplot)
library(FactoMineR)
library(dbplyr)
library(ROCR) 
library(ggridges) 
library(dendextend) 
library(e1071)
library(maps)
library(utils)
library(car)
library(factoextra)
library(Matrix)
library(arules)
library(cluster)
library(arulesViz)
#update.packages()
# set global chunk options: images will be 8x5 inches
knitr::opts_chunk$set(fig.width=8, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
```




```{r echo=FALSE}
fmrsq <- function(nam, data, i) {
  r.squared <- NULL
  fm        <- as.formula(paste0("`", nam[i], "` ", "~ ."))
  m1        <- summary(lm(fm, data = data))$r.squared
  1 - m1
}

viftol <- function(model) {
  m   <- as.data.frame(model.matrix(model))[, -1]
  nam <- names(m)
  p   <- length(model$coefficients) - 1
  tol <- c()

  for (i in seq_len(p)) {
    tol[i] <- fmrsq(nam, m, i)
  }
  vifs <- 1 / tol
  list(nam = names(m), tol = tol, vifs = vifs)
}

check_model <- function(model) {
  if (!all(class(model) == "lm")) {
    cat("Hmmm.. Looks like you have specified the wrong model. Follow the below steps to debug this error.
* Check if you have used the lm() function to build the model.
* If you have never used it before, you can learn more by typing ?lm or help(lm) in the Console\n.
Below is an example using mtcars data:\n lm(formula = mpg ~ disp + hp + wt, data = mtcars)\n",
"Happy modeling :)\n")
    stop("", call. = FALSE)
  }
}

check_logic <- function(logic) {
  lval        <- is.logical(logic)
  logic_class <- class(logic)
  logic_name  <- deparse(substitute(logic))

  if (lval != TRUE) {
    cat("\n *", logic_name, "can take only 2 values, either TRUE or FALSE", "\n * You have used", logic, "\n * Replace", logic, "with either TRUE or FALSE", "\n\n")
    stop("", call. = FALSE)
  }
}

ols_vif_tol <- function(model) {
  check_model(model)
  vt <- viftol(model)
  data.frame(Variables = vt$nam, Tolerance = vt$tol, VIF = vt$vifs)
}
```

[corr_simple()](https://towardsdatascience.com/how-to-create-a-correlation-matrix-with-too-many-variables-309cc0c0a57)
```{r echo=FALSE}
corr_simple <- function(data=df,sig=0.5){
  #convert data to numeric in order to run correlations
  #convert to factor first to keep the integrity of the data - each value will become a number rather than turn into NA
  df_cor <- data %>% mutate_if(is.character, as.factor)
  df_cor <- df_cor %>% mutate_if(is.factor, as.numeric)
  #run a correlation and drop the insignificant ones
  corr <- cor(df_cor)
  #prepare to drop duplicates and correlations of 1     
  corr[lower.tri(corr,diag=TRUE)] <- NA 
  #drop perfect correlations
  corr[corr == 1] <- NA 
  #turn into a 3-column table
  corr <- as.data.frame(as.table(corr))
  #remove the NA values from above 
  corr <- na.omit(corr) 
  #select significant values  
  corr <- subset(corr, abs(Freq) > sig) 
  #sort by highest correlation
  corr <- corr[order(-abs(corr$Freq)),] 
  #print table
  print(corr)
  #turn corr back into matrix in order to plot with corrplot
  mtx_corr <- reshape2::acast(corr, Var1~Var2, value.var="Freq")
  
  #plot correlations visually
  corrplot(mtx_corr, is.corr=FALSE, tl.col="black", na.label=" ")
}
```


# Instructions and Expectations


Predicting voter behavior is complicated for many reasons despite the tremendous effort in collecting, cleaning, analyzing, and understanding many available datasets. As the midterm elections in the United States is currently being held near the midpoint of the current president’s four-year term of office, it is a good time for us to review the 2020 United States presidential election data! Despite that the 2016 presidential election came as a big surprise to many, Biden’s victory in the 2020 presidential election has been widely predicted (e.g., see the well-known Nate Silver in FiveThirtyEight).


# Data

We will start the analysis with two data sets. The first one is the election data, which is drawn from here. The data contains county-level election results.

The second dataset is the 2017 United States county-level census data, which is available here.

The following code load in these two data sets: `election.raw` and `census.`

```{r message=FALSE, echo=FALSE}
## read data and convert candidate names and party names from string to factor
## we manually remove the variable "won", the indicator of county level winner
## In Problem 5 we will reproduce this variable!
election.raw <- read_csv("data/raw/candidates_county.csv", col_names = TRUE) %>% 
  mutate(candidate = as.factor(candidate), party = as.factor(party), won = NULL)


## remove the word "County" from the county names
words.to.remove = c("County")
remove.words <- function(str, words.to.remove){
  sapply(str, function(str){
    x <- unlist(strsplit(str, " "))
    x <- x[!x %in% words.to.remove]
    return(paste(x, collapse = " "))
  }, simplify = "array", USE.NAMES = FALSE)
}
election.raw$county <- remove.words(election.raw$county, words.to.remove)


## read census data
census <- read_csv("data/raw/census_county.csv") 
```

# Exploratory Data Analysis

## Election Data

**1.  (1 pts) Compute the total number of distinct values in state in election.raw to verify that the data contains all states and a federal district**

The following is the first few rows of the `election.raw` data.

```{r echo=FALSE}
head(election.raw)
```

Checking dimensions of `election.raw` data set.
```{r echo=FALSE}
dim(election.raw)
```

Checking for missing values in `election.raw` data set. 
```{r echo=FALSE}
ifelse(sum(is.na(election.raw)) == 0 , "There are no missing values", "There are missing values")
```

Checking unique vlaues for each column
```{r echo=FALSE}
sapply(election.raw, function(x) n_distinct(x))
```


```{r echo=FALSE}
unique_counties_per_state <- election.raw %>%
  distinct(state, county) %>%
  arrange(state, county)
```

## Census data

The following is the first few rows of the `census` data. The column names are all very self-explanatory:

```{r echo=FALSE}
head(census)
```

**2. (1 pts) Report the dimension of `census`. (1 pts) Are there missing values in the data set? (1 pts) Compute the total number of distinct values in `county` in `census`. (1 pts) Compare the values of total number of distinct county in `census` with that in `election.raw`. (1 pts) Comment on your findings.**

Checking dimensions of `census` data set.
```{r echo=FALSE}
dim(census)
```

Checking for missing values in `census` data set.
```{r echo=FALSE}
ifelse(sum(is.na(census)) == 0 , "There are no missing values", "There are missing values")
```


Which column has missing values?
```{r echo=FALSE}
column_missing <- colSums(is.na(census))
print(column_missing[column_missing > 0])
```



```{r echo=FALSE}
sapply(census, function(x) n_distinct(x))[3]
```
There are more distinct counties in `election.raw` than in `census`. Grouping them by state and county (some states have the same county names), `census` has around 3,220 counties, while `election.raw` has 4,633 counties. Upon closer examination of state and county, we see that some counties are named differently despite being from the same state county. Also, `census` has one more state than `election.raw` since `census` includes Puerto Rico.

```{r echo=FALSE}
unique_counties_per_state <- census %>%
  distinct(CountyId) %>%
  arrange(CountyId)
```

# Data Wrangling

**3. (4 pts) Construct aggregated data sets from election.raw data: i.e.,**

* Keep the county-level data as it is in election.raw.
* Create a state-level summary into a election.state.
* Create a federal-level summary into a election.total.

```{r echo=FALSE, cache=TRUE}
election.state <- aggregate(total_votes~candidate+county+state, election.raw,
                FUN = sum)
# election.state <- election.raw %>% group_by(candidate, county, state) %>% summarise(total_votes = sum(total_votes), .groups = 'drop')
head(election.state)
```

```{r echo=FALSE, cache=TRUE}
election.total <- aggregate(total_votes~candidate + state,election.raw,
                FUN = sum)
# election.raw %>% group_by(candidate) %>% summarise(total_votes = sum(total_votes), .groups = 'drop')
head(election.total)
```

**4. (1 pts) How many named presidential candidates were there in the 2020 election? (2 pts) Draw a bar chart of all votes received by each candidate. You can split this into multiple plots or may prefer to plot the results on a log scale. Either way, the results should be clear and legible! (For fun: spot Kanye West among the presidential candidates!)**

There were about `r n_distinct((election.total$candidate)) - 2` presidential candidates in the 2020 election, not counting write-ins and none of these candidates.
```{r echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# Exclude 'Write-in' and 'None of these candidates' from the count
named_candidates <- election.total %>% 
  filter(candidate != "Write-in" & candidate != "None of these candidates") %>% 
  distinct(candidate)

num_named_candidates <- n_distinct(named_candidates$candidate)

cat("There were", num_named_candidates, "named presidential candidates in the 2020 election.\n")

```

```{r echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
votes_per_candidate <- election.total %>%
  group_by(candidate) %>%
  summarise(total_votes = sum(total_votes)) %>%
  arrange(desc(total_votes))

head(votes_per_candidate)
```

```{r echo=FALSE}
library(ggplot2)
library(dplyr)

# Assuming votes_per_candidate is already created as mentioned in the previous steps
# and y_max is calculated.
y_max <- max(votes_per_candidate$total_votes)

# Split the dataset into chunks of 5 candidates each
chunk_size <- 5
splits <- split(votes_per_candidate, (seq(nrow(votes_per_candidate)) - 1) %/% chunk_size)

# Plot each chunk with the same y scale and adjust text settings
plot_list <- lapply(seq_along(splits), function(i) {
  chunk <- splits[[i]]
  ggplot(chunk, aes(x=reorder(candidate, -total_votes), y=total_votes)) +
    geom_bar(stat="identity") +
    scale_y_continuous(trans='log10', limits=c(1, y_max)) +  # set limits
    theme(axis.text.x = element_text(angle=90, hjust=1, size=10),  # x-axis text
          axis.text.y = element_text(size=10)) +  # y-axis text
    labs(x="Candidate", y="Total Votes (log scale)") +
    # ggtitle(paste("Plot", i)) +  # Add a title for each plot
    theme_minimal()
})

# Print each plot individually
for (plot in plot_list) {
  print(plot)
}
```

**5. (6 pts) Create data sets county.winner and state.winner by taking the candidate with the highest proportion of votes in both county level and state level. Hint: to create county.winner, start with election.raw, group by state and county, compute total votes, and pct = votes/total as the proportion of votes. Then choose the highest row using top_n (variable state.winner is similar).**


```{r echo=FALSE, error=FALSE, message=FALSE}
county.winner <- election.raw %>%
  mutate_at(vars(state, county), tolower) %>%
  mutate(county = gsub(" county| parish", "", county)) %>% # removes county or parish
  mutate(county = gsub(" cty| plt", "", county)) %>%
  mutate(county = gsub(" townships| townshps", "", county)) %>%
  mutate(county = gsub(" township| pt", "", county)) %>%
  mutate(county = gsub("[.]", "", county)) %>% # removes periods
  group_by(state, county) %>% # Some states have the same county name. This will differentiate between counties in different states
  mutate(total = sum(total_votes)) %>%
  mutate(pct = total_votes / total) %>%
  top_n(1,pct) %>%
  ungroup()

head(county.winner)
```



```{r echo=FALSE, error=FALSE, message=FALSE}
state.winner <- election.raw %>%
  mutate_at(vars(state, county), tolower) %>%
  group_by(state, candidate, party) %>%
  summarise(total_state_votes = sum(total_votes)) %>%
  ungroup() %>%
  group_by(state) %>%
  mutate(total_votes_state = sum(total_state_votes)) %>%
  mutate(pct = total_state_votes / total_votes_state) %>%
  top_n(1, pct) %>%
  ungroup() %>%
  select(state, candidate, party, total_state_votes, pct)

# election.raw %>% select(state, county) %>% distinct()

head(state.winner)
```

# Visualization

Visualization is crucial for gaining insight and intuition during data mining. We will map our data onto maps.

The R package ggplot2 can be used to draw maps. Consider the following code.

```{r  warning=FALSE, echo=FALSE}
states <- map_data("state")

ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  theme_void() + # removes 
  guides(fill=FALSE)  # color legend is unnecessary and takes too long
```

The variable states contain information to draw white polygons, and fill-colors are determined by region.

**6. (4 pts) Use similar code to above to draw county-level map by creating counties = map_data("county"). Color by county.**

```{r echo=FALSE}
counties <- map_data("county") # head(counties)

ggplot(data = counties) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  theme_void() +
  guides(fill=FALSE)  # color legend is unnecessary and takes too long
```

Now color the map by the winning candidate for each state. First, combine states variable and state.winner we created earlier using left_join(). Note that left_join() needs to match up values of states to join the tables. A call to left_join() takes all the values from the first table and looks for matches in the second table. If it finds a match, it adds the data from the second table; if not, it adds missing values:

Here, we’ll be combing the two data sets based on state name. However, the state names in states and state.winner can be in different formats: check them! Before using left_join(), use certain transform to make sure the state names in the two data sets: states (for map drawing) and state.winner (for coloring) are in the same formats. Then left_join(). Your figure will look similar to New York Times map.
.winner, state)

```{r echo=FALSE}
new_state <- states
# new_state$region <- str_to_title(new_state$region)

new_state <- new_state %>%
  left_join(state.winner, # # state.winner %>% mutate(state = tolower(state))
            by = c("region" = "state"))
# new_state %>% filter(is.na(region))
```


```{r echo=FALSE}
ggplot(data = new_state) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  theme_void() +
  guides()  # color legend is unnecessary and takes too long
```

**8. (6 pts) Color the map of the state of California by the winning candidate for each county.**


```{r echo=FALSE}
counties <- map_data("county")
new_counties <- counties

new_counties <- new_counties %>%
  left_join(county.winner,
            by = c("region" = "state", "subregion" = "county"))
```


```{r echo=FALSE}
# new_counties %>% filter(is.na(candidate))
california <- new_counties[which(new_counties$region == 'california'), ]
ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  theme_void() +
  guides()  + # color legend is unnecessary and takes too long
  ggtitle("Presidential Election in California")
```

**9. (4 pts) (Open-ended) Create a visualization of your choice using census data.** Many exit polls noted that demographics played a big role in the election. Use this Washington Post article and this R graph gallery for ideas and inspiration.


```{r echo=FALSE, message=FALSE, warnings=FALSE}
# Heat map
cali_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, alpha = pct, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  scale_alpha_continuous(range = c(0.4, 1), guide = FALSE) + # Adjust alpha range for visibility
  theme_minimal() +
  labs(fill = "Candidate", alpha = "Vote %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Heat Map of Candidates California Counties")

cali_map
```

```{r echo=FALSE}
# Combine data into a single data frame
# Ensure that `pct` is a continuous variable representing the percentage
california_combined <- california %>%
  mutate(candidate_pct = ifelse(candidate == "Joe Biden", pct, -pct))

# Create the plot
cali_map <- ggplot(california_combined, aes(x = long, y = lat, fill = candidate_pct, group = group)) + 
  geom_polygon(color = "white") + 
  coord_fixed(1.3) +
  scale_fill_gradientn(colours = c("red", "white", "blue"), 
                       values = scales::rescale(c(-1, 0, 1)), 
                       breaks = c(-1, 0, 1),
                       labels = c("Trump", "Neutral", "Biden"),
                       limits = c(-1, 1),
                       na.value = "transparent") +
  theme_minimal() +
  labs(fill = "Vote %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Heat Map of Candidates in California Counties")

# Print the plot
cali_map
```


```{r echo=FALSE}
library(patchwork)

counties <- map_data("county")

# temp <- census %>% mutate(County = as.character(gsub(" County", "", County)))
temp <- census %>% 
  mutate_at(vars(State, County), tolower) %>%
  mutate(County = gsub(" county|  parish", "", County))

temp_new_counties <- new_counties %>%
  left_join(temp, by = c("region" = "State", "subregion" = "County"))

# temp_new_counties <- new_counties

temp_new_counties <- temp_new_counties %>%
  mutate(majority_race = case_when(
    Hispanic > White & Hispanic > Black & Hispanic > Asian ~ "Hispanic",
    White > Hispanic & White > Black & White > Asian ~ "White",
    Black > Hispanic & Black > White & Black > Asian ~ "Black",
    Asian > Hispanic & Asian > White & Asian > Black ~ "Asian",
    TRUE ~ "Other"
  ))
```



```{r echo=FALSE}
california <- temp_new_counties[which(temp_new_counties$region == 'california'), ]

cali_white_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = White, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = '#E8E2CC', high = "#004434") + # scale_fill_gradient(low = '#E8E2CC', high = "#004434")
  theme_minimal() +
  labs(fill = "White %") +
  theme(text = element_text(size = 10)) +
  theme_void() + 
  ggtitle("White Percentage in California Counties")


cali_hisp_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = Hispanic, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "white", high = "orange") +
  theme_minimal() +
  labs(fill = "Hispanic %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Hispanic Percentage in California Counties")

cali_asian_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = Asian, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = 'white', high = "cornflowerblue") +
  theme_minimal() +
  labs(fill = "Asian %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Asian Percentage in California Counties")

cali_black_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = Black, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = 'white', high = "purple") + #scale_fill_gradient(low = '#DDEAD4', high = "#162D4A")
  theme_minimal() +
  labs(fill = "Black %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Black Percentage in California Counties")

```


```{r echo=FALSE}
cali_white_map <- cali_white_map + ggtitle("White Percentage")
cali_hisp_map <- cali_hisp_map + ggtitle("Hispanic Percentage")
cali_asian_map <- cali_asian_map + ggtitle("Asian Percentage")
cali_black_map <- cali_black_map + ggtitle("Black Percentage")
(cali_white_map + cali_hisp_map) / (cali_asian_map + cali_black_map)
```


```{r echo=FALSE}
cali_demog_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = majority_race, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Hispanic" = "orange", "White" = "green", "Black" = "purple", "Asian" = "cornflowerblue", "Other" = "gray")) +
  theme_minimal() +
  labs(fill = "Majority Race") +
  theme(text = element_text(size = 10)) + 
  theme_void() +
  ggtitle("Majority Demographic Group")

cali_map+cali_demog_map
```



```{r echo=FALSE}
texas <- temp_new_counties[which(temp_new_counties$region == 'texas'), ]
# candidate is missing for subregion "de witt"
# texas %>% filter(is.na(candidate))

texas_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, alpha = pct, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  scale_alpha_continuous(range = c(0.4, 1), guide = FALSE) + # Adjust alpha range for visibility
  theme_minimal() +
  labs(fill = "Candidate", alpha = "Vote %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Heat Map of Candidates Texas Counties")

texas_map
```


```{r echo=FALSE}
texas_white_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = White, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = '#E8E2CC', high = "#004434") + # scale_fill_gradient(low = '#E8E2CC', high = "#004434")
  theme_minimal() +
  labs(fill = "White %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("White Percentage in California Counties")

texas_hisp_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = Hispanic, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "white", high = "orange") +
  theme_minimal() +
  labs(fill = "Hispanic %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Hispanic Percentage in Texas Counties")

texas_asian_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = Asian, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = 'white', high = "cornflowerblue") +
  theme_minimal() +
  labs(fill = "Asian %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Asian Percentage in Texas Counties")

texas_black_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = Black, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = 'white', high = "purple") + #scale_fill_gradient(low = '#DDEAD4', high = "#162D4A")
  theme_minimal() +
  labs(fill = "Black %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Black Percentage in Texas Counties")

```


```{r echo=FALSE}
texas_white_map <- texas_white_map + ggtitle("White Percentage")
texas_hisp_map <- texas_hisp_map + ggtitle("Hispanic Percentage")
texas_asian_map <- texas_asian_map + ggtitle("Asian Percentage")
texas_black_map <- texas_black_map + ggtitle("Black Percentage")
(texas_white_map + texas_hisp_map)/ (texas_black_map + texas_asian_map)
```


```{r echo=FALSE}
texas <- temp_new_counties[which(temp_new_counties$region == 'texas'), ]

texas_demog_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = majority_race, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Hispanic" = "orange", "White" = "green", "Black" = "purple", "Asian" = "cyan", "Other" = "gray")) +
  theme_minimal() +
  labs(fill = "Majority Race") +
  theme(text = element_text(size = 10)) + theme_void() +
  ggtitle("Majority Demographic Group")

texas_map + texas_demog_map
```

# Data Cleaning

We start by cleaning the county-level census data through several transformations. Initially, we remove any rows with missing values using `na.omit()` and `filter()`. Then, we convert three attributes (Men, Employed, and VotingAgeCitizen) to percentages by dividing each by TotalPop and multiplying by 100. The code creates a new Minority attribute by combining Hispanic, Black, Native, Asian, and Pacific populations and removes these original variables after creating the Minority column using the `.keep = "unused" argument.` The Minority column is then relocated to appear after the White column for better organization.

The code removes several specified columns that are not needed for the analysis: IncomeErr, IncomePerCap, IncomePerCapErr, Walk, PublicWork, and Construction.

```{r}
census.clean <- census %>%
  na.omit() %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(Men = (Men/TotalPop) * 100) %>%
  mutate(Women = (Women/TotalPop) * 100) %>%
  mutate(Employed = (Employed/TotalPop) * 100) %>%
  mutate(VotingAgeCitizen = (VotingAgeCitizen/TotalPop) * 100) %>%
  mutate(Minority = Hispanic + Black + Native + Asian + Pacific, .keep = "unused") %>% # remove columns used to create Minority
  relocate(Minority, .after = White) %>%
  select(-c(IncomeErr, IncomePerCap, IncomePerCapErr, Walk, PublicWork, Construction))

head(census.clean, 5)
```


# Dimensionality reduction

After removing the State and County columns, the code creates a two-column data frame called pc.county containing the first two principal components (PC1 and PC2) from the PCA analysis of the cleaned census data.

```{r echo=FALSE}
temp_census.clean <- census.clean %>%
  select(-c(CountyId, State, County))
```

Regarding centering and scaling: The code uses both centering `(center=TRUE)` and scaling `(scale=TRUE)` before running PCA, which is appropriate because the variables in the census data have different scales. For example, some variables are percentages, while others might be raw counts or dollar values. Without scaling, variables with more significant variances would dominate the principal components, regardless of their importance.

```{r}
pr.out.census <- prcomp(temp_census.clean, scale=TRUE, center = TRUE)
```


```{r echo=FALSE}
# pc.county <- pr.out.census$x[,1:2]
pc.county <- data.frame(PC1 = pr.out.census$x[, 1], PC2 = pr.out.census$x[, 2])
```


The three features with the largest absolute values of the first principal component are:
```{r echo=FALSE}
new_p <- pr.out.census$rotation[,1]
largest_abs_loadings <- sort(abs(new_p), decreasing = TRUE)[1:3]
largest_features <- names(largest_abs_loadings)
```
`r names(largest_abs_loadings)`

Several features have negative values in the first principal component, including
```{r echo=FALSE}
negative_loadings <- new_p[new_p < 0]
```
`r names(negative_loadings)`


```{r message=FALSE, warnings=FALSE, results='hide', echo=FALSE, fig.height = 8, fig.width = 6}
# Extract loadings
loading_df <- data.frame(
  Taxon = colnames(temp_census.clean),  # Replace with the actual dataset column names
  PC1 = pr.out.census$rotation[, 1],      # First principal component
  PC2 = pr.out.census$rotation[, 2]       # Second principal component
)

loading_plot_df <- loading_df %>%
  pivot_longer(
    cols = starts_with("PC"),  # Columns to pivot (e.g., PC1, PC2)
    names_to = "Principal_Component",
    values_to = "Loading"
  )

# Base plot
base <- ggplot(loading_plot_df, aes(x = Loading, y = Taxon, color = Principal_Component)) +
  geom_point() +
  geom_line(aes(group = Taxon)) +  # Group lines by Taxon
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray", linewidth = .6) +  # Horizontal line at zero
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 7),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14)
  ) +
  labs(x = "Loading", y = "", color = "Principal Component")

fig7 <- base + facet_wrap(~Principal_Component, ncol = 2)

fig7
```

Notably, Poverty and ChildPoverty have negative signs and large magnitudes. Features with opposite signs in the principal component typically indicate a negative correlation between them. This means that as one feature increases, the other tends to decrease. In this case, the negative correlation suggests that counties with higher values in these socioeconomic indicators tend to have lower values in other related features.

For instance, counties with high poverty rates might be associated with lower income levels, fewer job opportunities, and different commuting patterns. The opposite signs in the first principal component capture this underlying relationship in the data, showing how different socioeconomic and demographic variables are interconnected.


```{r echo=FALSE}
##### **Combine PCA results with county and state**
# Extract the principal components
pca_data <- as.data.frame(pr.out.census$x)

# Add the county id, county, and state information back to the PCA results
pca_data$CountyId <- census.clean$CountyId
pca_data$County <- census.clean$County
pca_data$State <- census.clean$State
pca_data <- pca_data %>%
  select(CountyId, County, State, everything())

# head(pca_data)
```


## Determine the number of minimum number of PCs needed to capture 90% of the variance for the analysis

The code calculates the proportion of variance explained (PVE) for each principal component by:
1. Extracting the standard deviations from the principal component analysis
2. Calculating the variance by squaring the standard deviations
3. Computing the proportion of variance by dividing each component's variance by the total variance

```{r}
x <- pr.out.census$sdev
pr.var <- x^2
pve <- pr.var/sum(pr.var)
```

```{r echo=FALSE, message=FALSE, results='hide'}
# We want the cumulative at each index. The index is the PC. So index 1 is PC 1 
# and so on. We then want to know at which index is when it is greater than or equal to
# 90% ergo which(cumsum(pve) >= .9). Once we know which indexes fit this condition,
# we want to know the first index to fit this condition ergo min(...). 
# Therefore: 
# min(which(cumsum(pve) >= .9))
```
Computing `min(which(cumsum(pve) >= .9))`, we need about `r min(which(cumsum(pve) >= .9))` PCs in order to explain 90% of the total variation in the data. Thus, we need to retain the first `r min(which(cumsum(pve) >= .9))` components to explain at least 90% of the variability in the original data.


The cumulative proportion of variance plot visually demonstrates how the explained variance accumulates as more principal components are included, allowing us to see the incremental contribution of each additional component to the total variance explained.

```{r echo=FALSE}
plot(cumsum(pve), xlab="Principal Component ",
     ylab=" Cumulative Proportion of Variance Explained ", ylim=c(0,1), type='b')
```


# Clustering

```{r echo=FALSE}
census.clean <- census %>%
  na.omit() %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(Men = (Men/TotalPop) * 100) %>%
  mutate(Women = (Women/TotalPop) * 100) %>%
  mutate(Employed = (Employed/TotalPop) * 100) %>%
  mutate(VotingAgeCitizen = (VotingAgeCitizen/TotalPop) * 100) %>%
  mutate(Minority = Hispanic + Black + Native + Asian + Pacific, .keep = "unused") %>%
  relocate(Minority, .after = White) %>%
  select(-c(IncomeErr, IncomePerCap, IncomePerCapErr, Walk, PublicWork, Construction))
```

### Centering and Scaling
Same approach as dimensionality reduction.
```{r}
scar <- scale(census.clean[,c(-1:-3)], center=TRUE, scale=TRUE)
census.clean.dist <- dist(scar)
set.seed(10)
census.clean.hclust <- hclust(census.clean.dist)
```


```{r echo=FALSE}
set.seed(10)
# Get cluster
clus <- cutree(census.clean.hclust, 10)
```

### First Clustering (Original Features):
When performing hierarchical clustering using the scaled original features with complete linkage and cutting the tree into 10 clusters, the distribution of observations across clusters is quite uneven. Cluster 1 contains the most observations, with 2,967 counties, while several other clusters have very few counties. For instance, clusters 9 and 10 have only 4 and 1 counties, respectively, indicating that the clustering algorithm identified a few very distinct groups among the counties.
```{r echo=FALSE}
table_data <- as.data.frame(table(clus))

table(clus)
```


```{r echo=FALSE}
census_transform <- as.data.frame(pc.county)

census_transform.dist <- dist(pc.county)

set.seed(10)
census_transform.hclust <- hclust(census_transform.dist)
clus_new <- cutree(census_transform.hclust, 10)
```


### Second Clustering (First Two Principal Components):
The cluster distribution changes significantly after re-running the hierarchical clustering using the first two principal components from pc.county. Cluster 1 contains 1,670 counties, cluster 7 has 648 counties, and Cluster 2 has 563 counties. The remaining clusters have fewer counties, but the distribution differs from the first clustering.
```{r echo=FALSE}
table_data <- as.data.frame(table(clus_new))

table(clus_new)
```

What can we take out of this

* Both clustering approaches result in highly imbalanced cluster sizes, suggesting some very distinct groups of counties.
* The uneven cluster sizes in both methods suggest that a few very distinct county groups stand out from the majority, which could represent counties with unique demographic or economic characteristics.



```{r echo=FALSE}
### The Elbow Method 
# Calculate scaled data for clustering
scaled_data <- scale(census.clean[,c(-1:-3)], center=TRUE, scale=TRUE)

# Calculate distances and perform hierarchical clustering
distance_matrix <- dist(pc.county)
hclust_result <- hclust(distance_matrix)

# Elbow Method for Optimal Number of Clusters
wss <- sapply(2:15, function(k){
  clusters <- cutree(hclust_result, k)
  wss <- sum(kmeans(scaled_data, centers = k)$withinss)
  return(wss)
  }
)

# Prepare data for ggplot
elbow_data <- data.frame(k = 2:15, wss = wss)

# Plotting WSS vs. Number of Clusters
elbow_method <- ggplot(elbow_data, aes(x = k, y = wss)) +
  geom_point() +
  geom_line() +
  labs(title = "Elbow Method",
       x = "Number of Clusters (k)",
       y = "WCSS") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```



# Classification

We exclude the predictor party from election.cl because the classification task aims to predict a county's winner using census information. Including party as a predictor would create data leakage, as it directly informs us about the candidate's party affiliation rather than the demographic or socioeconomic factors of the county.

The party variable is tied to the outcome (the winner), and its inclusion would undermine the model's ability to generalize by basing predictions on the relationship between party and candidate rather than the census data. We remove the party variable from the dataset to ensure the model learns from relevant predictors.

```{r}
# we move all state and county names into lower-case
tmpwinner <- county.winner %>% ungroup %>%
  mutate_at(vars(state, county), tolower)

# we move all state and county names into lower-case
# we further remove suffixes of "county" and "parish"
tmpcensus <- census.clean %>% mutate_at(vars(State, County), tolower) %>%
  mutate(County = gsub(" county|  parish", "", County))

# we join the two datasets
election.cl <- tmpwinner %>%
  left_join(tmpcensus, by = c("state"="State", "county"="County")) %>%
  na.omit

# drop levels of county winners if you haven't done so in previous parts
election.cl$candidate <- droplevels(election.cl$candidate)

## save meta information
election.meta <- election.cl %>% select(c(county, party, CountyId, state, total_votes, pct, total))

## save predictors and class labels
election.cl = election.cl %>% select(-c(county, party, CountyId, state, total_votes, pct, total))
```


# Classification

Using the following code, partition data into 80% training and 20% testing:

```{r}
set.seed(10)
n <- nrow(election.cl)
idx.tr <- sample.int(n, 0.8*n)
election.tr <- election.cl[idx.tr, ]
election.te <- election.cl[-idx.tr, ]
```

Use the following code to define 10 cross-validation folds:

```{r echo=FALSE}
set.seed(10)
nfold <- 10
folds <- sample(cut(1:nrow(election.tr), breaks=nfold, labels=FALSE))
```

Using the following error rate function. And the object records is used to record the classification performance of each method in the subsequent problems.

```{r echo=FALSE}
calc_error_rate = function(predicted.value, true.value){
  return(mean(true.value!=predicted.value))
}
records = matrix(NA, nrow=6, ncol=2)
colnames(records) = c("train.error","test.error")
rownames(records) = c("Decision Tree","Logistic Regression","Lasso Logistic Regression","Random Forest","Neural Network", "Neural Network Optimized")
```


**15. Decision tree: (2 pts) train a decision tree by cv.tree().** (2 pts) Prune tree to minimize misclassification error. Be sure to use the folds from above for cross-validation. (2 pts) Visualize the trees before and after pruning. (1 pts) Save training and test errors to records object. (2 pts) Interpret and discuss the results of the decision tree analysis. (2 pts) Use this plot to tell a story about voting behavior.

Decision Tree purity

```{r echo=FALSE}
tree.election <- tree(candidate~.,data = election.tr)
```

```{r echo=FALSE}
draw.tree(tree.election, nodeinfo=TRUE, cex = 0.5)
title("Single Tree Built on Training Set")
```

```{r echo=FALSE}
set.seed(20)
cv.election <- cv.tree(tree.election, FUN = prune.misclass, rand = folds)
best_size <- cv.election$size[which.min(cv.election$dev)]
```

From our function `cv.tree()`, the best size is `r best_size`.

```{r}
cv_data <- data.frame(
  size = cv.election$size,
  error = cv.election$dev
)

# Create the plot
ggplot(cv_data, aes(x = size, y = error)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Cross-Validation Error vs Tree Size",
    x = "Tree Size",
    y = "Cross-Validation Error"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = unique(cv_data$size)) +  # All tree sizes on x-axis
  geom_point(
    data = cv_data[which.min(cv_data$error), ],
    aes(x = size, y = error),
    color = "red",
    size = 4
  )  # Highlight minimum error point
```

```{r echo=FALSE}
# Prune Tree
pt.cv <- prune.misclass(tree.election, best = best_size)
```

```{r echo=FALSE}
# Plot Pruned Tree
draw.tree(pt.cv, nodeinfo=TRUE, cex = 0.8)
title(paste("Pruned Tree of Size ", best_size))
```



**Both trees have Transit first followed by White and Women. Before pruning, White was right below Women, but in after pruning, White is at the very bottom of Women. Minority is still at the same place. In after pruning, Professional is not in the White split, but Professional is still in the Women split. Women is absent in the White split. Men is also absent in the White split.**

**It seems White was a big contributing factor if you were going to vote for Trump, while it seems Women was a big contributing factor if you were going to vote for Biden.**

```{r echo=FALSE}
# Predict on test set
pred.train.tree = predict(pt.cv, election.tr, type="class")
pred.test.tree = predict(pt.cv, election.te, type="class")

# Obtain confusion matrix
err.pt.cv <- table(pred.test.tree, election.te$candidate) # table(pred.train.tree, election.tr$candidate)
err.pt.cv
```

```{r echo=FALSE}
a <- calc_error_rate(pred.train.tree, election.tr$candidate)
b <- calc_error_rate(pred.test.tree, election.te$candidate)
records[1,] <- c(a,b)
```

**16. (2 pts) Run a logistic regression to predict the winning candidate in each county.** (1 pts) Save training and test errors to records variable. (1 pts) What are the significant variables? (1 pts) Are they consistent with what you saw in decision tree analysis? (2 pts) Interpret the meaning of a couple of the significant coefficients in terms of a unit change in the variables.


For interpretable machine learning
```{r echo=FALSE}
# Fitting model using logistic regression
glm.obj.election <- glm(candidate~.,data = election.tr, family = binomial)
```

```{r echo=FALSE}
set.seed(20) 
prob.training <- predict(glm.obj.election, election.tr, type="response")
prob.test <- predict(glm.obj.election, election.te, type="response")

# Convert probabilities to binary predictions
majority_rule <- 0.5
pred_training <- ifelse(prob.training > majority_rule, "Joe Biden", "Donald Trump")
pred_test <- ifelse(prob.test > majority_rule, "Joe Biden", "Donald Trump")
```

```{r echo=FALSE}
training_pred_table <- table(Predicted = pred_training, True = election.tr$candidate)
test_pred_table <- table(Predicted = pred_test, True = election.te$candidate)

train_accuracy <- sum(diag(training_pred_table)) / nrow(election.tr)
test_accuracy <- sum(diag(test_pred_table)) / nrow(election.te)
```

```{r echo=FALSE}
c <- calc_error_rate(pred_training, election.tr$candidate)
d <- calc_error_rate(pred_test, election.te$candidate)
records[2,] <- c(c,d)
```

```{r echo=FALSE}
# summary(glm.obj.election)$coefficients
```

Below are the significant variables:
```{r echo=FALSE}
tail(sort(abs(glm.obj.election$coefficients[2:24])), 5)
```
White is included, but Women and Minority are not among the top 5. It does deviate from the Decision Tree analysis.

If were to increase Production by one unit, holding the rest fixed, then applying the same logic to Professional, Employed, Service, and Unemployment, we would get percent in the odds:
```{r echo=FALSE}
(exp(tail(sort(glm.obj.election$coefficients[2:24]), 5)) - 1)*100
```

**17. You may notice that you get a warning glm.fit: fitted probabilities numerically 0 or 1 occurred.** As we discussed in class, this is an indication that we have perfect separation (some linear combination of variables perfectly predicts the winner). This is usually a sign that we are overfitting. One way to control overfitting in logistic regression is through regularization.

(3 pts) Use the cv.glmnet function from the glmnet library to run a 10-fold cross validation and select the best regularization parameter for the logistic regression with LASSO penalty. Set lambda = seq(1, 50) * 1e-4 in cv.glmnet() function to set pre-defined candidate values for the tuning parameter λ.

```{r echo=FALSE}
set.seed(10)
dat <- model.matrix(candidate~., election.cl) 
idx.tr <- sample.int(n, 0.8*n)
x.train = dat[idx.tr, ] 
y.train = election.cl$candidate[idx.tr]

# The rest as test data 
x.test = dat[-idx.tr, ] 
y.test <- election.cl$candidate[-idx.tr]
```




```{r echo=FALSE}
set.seed(10)
cv.out.lasso <- cv.glmnet(x.train, y.train, 
                          nfolds = 10, alpha = 1, lambda = (1e-4) * seq(1, 50),
                          family = "binomial")
bestlam <- cv.out.lasso$lambda.min
```

(1 pts) What is the optimal value of λ in cross validation? (1 pts) What are the non-zero coefficients in the LASSO regression for the optimal value of λ? (1 pts) How do they compare to the unpenalized logistic regression? (1 pts) Comment on the comparison. (1 pts) Save training and test errors to the records variable.

The optimal value of $\lambda$ in cross validation is `r bestlam`.

Below are the non-zero coefficients in the LASSO regression for the optimal value of λ:
```{r echo=FALSE}
lasso.model <- glmnet(x.train, y.train, alpha=1, family = "binomial")
pred <- predict(lasso.model,type="coefficients",s=bestlam)[1:24,]
```

```{r echo=FALSE}
dense_coefficients <- as.matrix(coef(lasso.model, s = bestlam))
non_zero_coefficients <- dense_coefficients[dense_coefficients != 0, ]
print(non_zero_coefficients)
```

```{r}
set.seed(10)

# Predict probabilities for training and test datasets
prob.training <- predict(lasso.model, type = "response", s = bestlam, newx = x.train)
prob.test <- predict(lasso.model, type = "response", s = bestlam, newx = x.test)

# Generate predictions using majority rule
majority_rule <- 0.5
pred_training <- ifelse(prob.training > majority_rule, "Joe Biden", "Donald Trump")
pred_test <- ifelse(prob.test > majority_rule, "Joe Biden", "Donald Trump")
```


```{r}
# Training set confusion matrix and error calculation
training_pred_table <- table(Predicted = pred_training, True = election.tr$candidate)
percent_train <- sum(diag(training_pred_table)) / nrow(election.tr) * 100

# Test set confusion matrix and error calculation
test_pred_table <- table(Predicted = pred_test, True = election.te$candidate)
percent_test <- sum(diag(test_pred_table)) / nrow(election.te) * 100
```


```{r echo=FALSE}
e <- calc_error_rate(as.factor(pred_training), election.tr$candidate)
f <- calc_error_rate(as.factor(pred_test), election.te$candidate)
records[3,] <- c(e,f)
```

Unpenalized logistic regression: When taking their absolute values, Women and Income had low coefficients compared to the rest of the variables. Excluding these two, the other variables were within range of each other.

LASSO regression: Minority, Income, WorkAtHome, and MeanCommute had zero coefficients.

It does seem Income is not an important metric. White has been consistently shown as being an important variable.


**18. (6 pts) Compute ROC curves for the decision tree, logistic regression and LASSO logistic regression using predictions on the test data.** Display them on the same plot. (2 pts) Based on your classification results, discuss the pros and cons of the various methods. (2 pts) Are the different classifiers more appropriate for answering different kinds of questions about the election?

```{r echo=FALSE}
# Decision Tree - get the probabilities for "Joe Biden"
pred_tree <- predict(pt.cv, election.te, type="vector")[, 2]

# Logistic Regression
pred_log <- predict(glm.obj.election, election.te, type="response")

# LASS Logistic Regression
pred_lasso <- predict(lasso.model, type="response", s=bestlam, newx = x.test)
```


```{r echo=FALSE}
roc_pred_tree <-
   prediction(
     predictions = pred_tree,
     labels = election.te$candidate
)

roc_pred_log <-
   prediction(
     predictions = pred_log,
     labels = election.te$candidate
)

roc_pred_lasso <-
   prediction(
     predictions = pred_lasso,
     labels = y.test
)
```


```{r echo=FALSE}
roc_perf_tree <- performance(roc_pred_tree, measure = "tpr", x.measure = "fpr")
roc_perf_log <- performance(roc_pred_log, measure = "tpr", x.measure = "fpr")
roc_perf_lasso <- performance(roc_pred_lasso, measure = "tpr", x.measure = "fpr")
```

```{r before-random, echo=FALSE}
roc_data <- data.frame(
  FPR = c(
    unlist(roc_perf_tree@x.values), 
    unlist(roc_perf_log@x.values), 
    unlist(roc_perf_lasso@x.values)),
  TPR = c(
    unlist(roc_perf_tree@y.values), 
    unlist(roc_perf_log@y.values), 
    unlist(roc_perf_lasso@y.values)),
  Model = factor(rep(c("Tree", "Log", "Lasso"), times = c(length(roc_perf_tree@x.values[[1]]),
                                                          length(roc_perf_log@x.values[[1]]),
                                                          length(roc_perf_lasso@x.values[[1]])
                                                          )
                     )
                 )
)

ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("green", "blue", "red")) +
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", # (FPR)
       y = "True Positive Rate", # (TPR)
       color = "Model",
       linetype = "Model") +
  #theme_minimal() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") # Positions legend at the bottom
```


**Answer:**

**Answer: **

# Taking it further

This part will be worth up to a 20% of your final project grade!

**19. (9 pts) Explore additional classification methods.** Consider applying additional two classification methods from KNN, LDA, QDA, SVM, random forest, boosting, neural networks etc. (You may research and use methods beyond those covered in this course). How do these compare to the tree method, logistic regression, and the lasso logistic regression?

### Random Forest

```{r echo=FALSE, results='hide'}
election.randfor <- randomForest(candidate~., 
                                 data = election.tr, importance=TRUE)
```

```{r echo=FALSE}
pred_training <- predict(election.randfor, newdata = election.tr, type = "response")
pred_test <- predict(election.randfor, newdata = election.te, type = "response")

a <- calc_error_rate(as.factor(pred_training), election.tr$candidate)
b <- calc_error_rate(as.factor(pred_test), election.te$candidate)
records[4,] <- c(a,b)
```

```{r echo=FALSE}
pred_randfor <- predict(election.randfor, newdata = election.te, type = "prob")[, 2]

roc_pred_randfor <-
   prediction(
     predictions = pred_randfor,
     labels = election.te$candidate
)

roc_perf_randfor <- performance(roc_pred_randfor, measure = "tpr", x.measure = "fpr")
```


### Neural Net

```{r echo=FALSE}
set.seed(10)
n <- nrow(election.cl)
idx.tr <- sample.int(n, 0.8*n)
election.tr <- election.cl[idx.tr, ]
election.te <- election.cl[-idx.tr, ]
```


```{r echo=FALSE}
# Load required libraries
library(nnet)
library(caret)  # For data preprocessing
library(parallel)
library(doParallel)
library(parallelly)

# Scale our predictors for neural network
# Create preprocessing object using training data
preprocess_params <- preProcess(election.tr[,-1], method = c("center", "scale"))

# Apply preprocessing to both training and test sets
X_train_scaled <- predict(preprocess_params, election.tr[,-1])
X_test_scaled <- predict(preprocess_params, election.te[,-1])

# Add the response variable back
train_data_scaled <- data.frame(
  candidate = election.tr$candidate,
  X_train_scaled
)
test_data_scaled <- data.frame(
  candidate = election.te$candidate,
  X_test_scaled
)

# Set seed for reproducibility
set.seed(123)

# Train neural network
nn_model <- nnet(
  candidate ~ ., 
  data = train_data_scaled,
  size = 5,        # number of hidden nodes
  decay = 0.1,     # weight decay for regularization
  maxit = 1000,    # maximum iterations
  trace = FALSE    # suppress training output
)


# Make predictions
train_pred_prob <- predict(nn_model, train_data_scaled, type = "raw")
test_pred_prob <- predict(nn_model, test_data_scaled, type = "raw")

# Convert probabilities to class predictions
train_pred <- ifelse(train_pred_prob > 0.5, "Joe Biden", "Donald Trump")
test_pred <- ifelse(test_pred_prob > 0.5, "Joe Biden", "Donald Trump")
```

```{r echo=FALSE}
# Storing Error Rates
a <- calc_error_rate(as.factor(train_pred), election.tr$candidate)
b <- calc_error_rate(as.factor(test_pred), election.te$candidate)
records[5,] <- c(a,b)

roc_pred_nn <- prediction(
    predictions = test_pred_prob,
    labels = election.te$candidate #as.factor(election.te$candidate, levels = c("Donald Trump", "Joe Biden"))
)

roc_perf_nn <- performance(roc_pred_nn, measure = "tpr", x.measure = "fpr")
```


```{r nn-opt, echo=FALSE}
# Modify factor levels to be valid R names
election.tr$candidate <- factor(election.tr$candidate,
                              levels = c("Donald Trump", "Joe Biden"),
                              labels = c("Donald_Trump", "Joe_Biden"))
election.te$candidate <- factor(election.te$candidate,
                              levels = c("Donald Trump", "Joe Biden"),
                              labels = c("Donald_Trump", "Joe_Biden"))

# Setup parallel processing
num_cores <- availableCores(omit = 1)  # Leave one core free
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Create training control object
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  verboseIter = TRUE,
  allowParallel = TRUE
)

# Define grid of hyperparameters
grid <- expand.grid(
  size = c(3, 5, 7, 10, 13),
  decay = c(0.001, 0.01, 0.1, 0.3, 0.5)
)

# Set seed
set.seed(10)

# Train model with grid search
nnet_optimal <- train(
  x = X_train_scaled,
  y = election.tr$candidate,
  method = "nnet",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = grid,
  trace = FALSE,
  maxit = 1000,
  linout = FALSE,
  MaxNWts = 1000
)

# Stop parallel processing
stopCluster(cl)

# Make predictions with optimal model
train_pred_prob <- predict(nnet_optimal, X_train_scaled, type = "prob")[,"Joe_Biden"]
test_pred_prob <- predict(nnet_optimal, X_test_scaled, type = "prob")[,"Joe_Biden"]

# Calculate error rates
train_pred <- ifelse(train_pred_prob > 0.5, "Joe_Biden", "Donald_Trump")
test_pred <- ifelse(test_pred_prob > 0.5, "Joe_Biden", "Donald_Trump")
```

```{r echo=FALSE}
# Plot tuning results
plot(nnet_optimal)
```

```{r echo=FALSE}
# Storing Error Rates
a <- calc_error_rate(as.factor(train_pred), election.tr$candidate)
b <- calc_error_rate(as.factor(test_pred), election.te$candidate)
records[6,] <- c(a,b)


roc_pred_opt <- prediction(
    predictions = test_pred_prob,
    labels = factor(election.te$candidate, levels = c("Donald_Trump", "Joe_Biden"))
)

roc_perf_opt <- performance(roc_pred_opt, measure = "tpr", x.measure = "fpr")
```


```{r echo=FALSE}
roc_data <- data.frame(
  FPR = c(
    unlist(roc_perf_randfor@x.values), # Random Forest
    unlist(roc_perf_nn@x.values), # Neural Net
    unlist(roc_perf_opt@x.values) # Neural Net Optimized
    ),
  TPR = c(
    unlist(roc_perf_randfor@y.values), # Random Forest
    unlist(roc_perf_nn@y.values), # Neural Net
    unlist(roc_perf_opt@y.values) # Neural Net Optimized
    ),
  Model = factor(rep(c("Random Forest", "Neural Net", "Neural Net Optimized"), 
                     times = c(length(roc_perf_randfor@x.values[[1]]),
                               length(roc_perf_nn@x.values[[1]]),
                               length(roc_perf_opt@x.values[[1]])
                               )
                     )
                 )
)

ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("purple", "brown4", "cadetblue")) +
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", # (FPR)
       y = "True Positive Rate", # (TPR)
       color = "Model",
       linetype = "Model") +
  #theme_minimal() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") # Positions legend at the bottom
```






```{r echo=FALSE}
roc_data <- data.frame(
  FPR = c(
    unlist(roc_perf_randfor@x.values), # Random Forest
    unlist(roc_perf_lasso@x.values), # LASSO Logistic Regression
    unlist(roc_perf_opt@x.values) # Neural Net Optimized
    ),
  TPR = c(
    unlist(roc_perf_randfor@y.values), # Random Forest
    unlist(roc_perf_lasso@y.values), # LASSO Logistic Regression
    unlist(roc_perf_opt@y.values) # Neural Net Optimized
    ),
  Model = factor(rep(c("Random Forest", "LASSO", "Neural Net Optimized"), 
                     times = c(length(roc_perf_randfor@x.values[[1]]),
                               length(roc_perf_lasso@x.values[[1]]),
                               length(roc_perf_opt@x.values[[1]])
                               )
                     )
                 )
)

ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("red", "brown4", "cadetblue")) +
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", # (FPR)
       y = "True Positive Rate", # (TPR)
       color = "Model",
       linetype = "Model") +
  #theme_minimal() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") # Positions legend at the bottom
```

**Answer:**

**Plotting the ROC Curve for all the methods, we see that our CV Decision Tree performs the worst, then Random Forest. After that, it's a matter of personal preference. Random Forest does outperform CV Decision Tree throughout the ROC Curve. In my opinion Decision Tree and Random Forest would not be suitable for our problem since we already narrowed it down to two candidates. I still think LASSO Logistic Regression is the most optimal, followed by our regular Logistic Regression.**
