---
title: "Exploratory Analysis of the 2020 Presidential Election"
author: "Alexander Sanchez"
output: github_document
---


```{r setup, echo=FALSE, include=FALSE}
# install.packages("arulesViz")
library(knitr)
library(ISLR) 
library(tree)
library(maptree)
library(glmnet) 
library(caret)
library(psych)
library(tidyverse)
library(corrplot)
library(FactoMineR)
library(dbplyr)
library(ROCR) 
library(ggridges) 
library(dendextend) 
library(e1071)
library(maps)
library(utils)
library(car)
library(factoextra)
library(Matrix)
library(arules)
library(arulesViz)
library(ggplot2)
library(dplyr)
library(patchwork)
#update.packages()
# set global chunk options: images will be 8x5 inches
knitr::opts_chunk$set(fig.width=8, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
```



# Instructions and Expectations


Predicting voter behavior is complicated for many reasons despite the tremendous effort in collecting, cleaning, analyzing, and understanding many available datasets. As the midterm elections in the United States is currently being held near the midpoint of the current president’s four-year term of office, it is a good time for us to review the 2020 United States presidential election data! Despite that the 2016 presidential election came as a big surprise to many, Biden’s victory in the 2020 presidential election has been widely predicted (e.g., see the well-known Nate Silver in FiveThirtyEight).


# Data

We will start the analysis with two data sets. The first one is the election data, which is drawn from here. The data contains county-level election results.

The second dataset is the 2017 United States county-level census data, which is available here.

The following code load in these two data sets: `election.raw` and `census.`

```{r message=FALSE, echo=FALSE}
## read data and convert candidate names and party names from string to factor
## we manually remove the variable "won", the indicator of county level winner
## In Problem 5 we will reproduce this variable!
election.raw <- read_csv("../data/raw/candidates_county.csv", col_names = TRUE) %>% 
  mutate(candidate = as.factor(candidate), party = as.factor(party), won = NULL)

## remove the word "County" from the county names
words.to.remove = c("County")
remove.words <- function(str, words.to.remove){
  sapply(str, function(str){
    x <- unlist(strsplit(str, " "))
    x <- x[!x %in% words.to.remove]
    return(paste(x, collapse = " "))
  }, simplify = "array", USE.NAMES = FALSE)
}
election.raw$county <- remove.words(election.raw$county, words.to.remove)

## read census data
census <- read_csv("../data/raw/census_county.csv") 
```

# Exploratory Data Analysis

## Election Data

**Before we perform any data visualization, we will analyze the election.raw dataset, including examining the data structure and ensuring that it includes all states and a federal district.**

The following is the first few rows of the `election.raw` data.

```{r echo=FALSE}
head(election.raw)
```

Checking dimensions of `election.raw` data set.
```{r echo=FALSE}
dim(election.raw)
```

Checking for missing values in `election.raw` data set. 
```{r echo=FALSE}
ifelse(sum(is.na(election.raw)) == 0 , "There are no missing values", "There are missing values")
```

Ensuring that it includes all states and a federal district
```{r echo=FALSE}
# Count distinct states
num_states <- n_distinct(election.raw$state)
print(paste("Number of distinct states/districts:", num_states))

# Verify equals 51 (not forgetting DC)
ifelse(num_states == 51, 
       "Confirmed: Data contains all 50 states and federal district", 
       paste("Issue: Expected 51, but found", num_states))
```


```{r echo=FALSE}
election_unique_counties_per_state <- election.raw %>%
  distinct(state, county) %>%
  arrange(state, county)
```

## Census data

The following is the first few rows of the `census` data. The column names are all very self-explanatory:

```{r echo=FALSE}
head(census)
```

### Census Data Exploration and Comparison with Election Data

**Next, we will examine the structure and completeness of the census dataset, identify any missing values, and compare the geographic coverage between the census and election datasets to understand potential data integration challenges.**

Checking dimensions of `census` data set.
```{r echo=FALSE}
dim(census)
```

Checking for missing values in `census` data set.
```{r echo=FALSE}
ifelse(sum(is.na(census)) == 0 , "There are no missing values", "There are missing values")
```


Which column has missing values?
```{r echo=FALSE}
column_missing <- colSums(is.na(census))
print(column_missing[column_missing > 0])
```


```{r echo=FALSE}
census_unique_counties_per_state <- census %>%
  distinct(State, County) %>%
  arrange(State, County)
```

There are more distinct counties in `election.raw` than in `census.` Grouping them by state and county (some states have the same county names), `census` has around `r nrow(census_unique_counties_per_state)` counties, while `election.raw` has `r nrow(election_unique_counties_per_state)` counties. Additionally, `census` has one more state than `election.raw`, as `census` includes Puerto Rico. Examining both datasets reveals that counties have different naming conventions despite referring to the same geographic areas. For example, Alabama appears as 'Autauga County' in `census` but 'Autauga' in `election.raw.` This naming inconsistency will need to be resolved when merging the datasets.

# Data Wrangling

**3. (4 pts) Construct aggregated data sets from election.raw data: i.e.,**

* Keep the county-level data as it is in election.raw.
* Create a state-level summary into a election.state.
* Create a federal-level summary into a election.total.

```{r echo=FALSE, cache=TRUE}
# State-level summary
election.state <- election.raw %>% 
  group_by(candidate, state) %>% 
  summarise(total_votes = sum(total_votes), .groups = 'drop')
# election.state <- election.raw %>% group_by(candidate, county, state) %>% summarise(total_votes = sum(total_votes), .groups = 'drop')
head(election.state)
```

```{r echo=FALSE, cache=TRUE}
# Federal-level summary  
election.total <- election.raw %>% 
  group_by(candidate) %>% 
  summarise(total_votes = sum(total_votes), .groups = 'drop')

head(election.total)
```


```{r echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
# Exclude 'Write-in' and 'None of these candidates' from the count
named_candidates <- election.total %>% 
  filter(candidate != "Write-ins" & candidate != "None of these candidates") %>% 
  distinct(candidate)

num_named_candidates <- n_distinct(named_candidates$candidate)
```
The 2020 presidential election featured `r num_named_candidates` named candidates on ballots across the United States, excluding "Write-ins" votes and "None of these candidates" options. This analysis examines the vote distribution among all candidates to understand the electoral landscape.

```{r echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
votes_per_candidate <- election.total %>%
  arrange(desc(total_votes))

# head(votes_per_candidate)
```

```{r echo=FALSE}
votes_per_candidate <- votes_per_candidate %>%
  mutate(party = case_when(
    candidate %in% c("Joe Biden") ~ "Democratic",
    candidate %in% c("Donald Trump") ~ "Republican", 
    TRUE ~ "Other"
  )) %>%
  head(10)
```


```{r echo=FALSE}
# Top 2 candidates and rest
top2 <- votes_per_candidate %>% head(2)
others <- votes_per_candidate %>% slice(3:10)

# Plotting the Major Party Candidates
p1 <- ggplot(top2, aes(x=total_votes, y=reorder(candidate, total_votes), fill=party)) +
  geom_bar(stat="identity", width=0.6) +
  geom_text(aes(label = paste(round(total_votes/1e6, 1), "M")), 
            hjust = 1.1, size = 4, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("Democratic" = "#0015BC", "Republican" = "#E9141D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 11),
    legend.position = "top",
    legend.title = element_blank()
  ) +
  labs(title = "Major Party Candidates", y = "")

# Plotting the Other Candidates
p2 <- ggplot(others, aes(x=total_votes, y=reorder(candidate, total_votes))) +
  geom_bar(stat="identity", width=0.6, fill="cadetblue") +
  geom_text(aes(label = paste(round(total_votes/1e3, 0), "K")), 
            hjust = 1.1, size = 3, fontface = "bold", color = "white") +
  scale_x_log10() +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 10)
  ) +
  labs(title = "Other Candidates", y = "")

# Combine plots
p1 / p2
```

The vote totals reveal a stark bifurcation in the American electoral system. As shown in the visualization below, the two major party candidates—Joe Biden (Democratic) and Donald Trump (Republican)—dominated the election, receiving over 74 million votes each. In contrast, all other candidates combined received significantly fewer votes, with the highest third-party candidate earning under 2 million votes.

This dramatic difference in vote scale necessitates a two-panel visualization approach. The upper panel displays the major party candidates using a linear scale, clearly showing Biden's victory margin of approximately 7 million votes. The lower panel focuses on third-party and independent candidates, using a logarithmic scale to facilitate meaningful comparisons within this group.

The data illustrates how the American two-party system creates substantial barriers for third-party candidates, regardless of their public profile.

**5. (6 pts) Create data sets county.winner and state.winner by taking the candidate with the highest proportion of votes in both county level and state level. Hint: to create county.winner, start with election.raw, group by state and county, compute total votes, and pct = votes/total as the proportion of votes. Then choose the highest row using top_n (variable state.winner is similar).**


```{r echo=FALSE, error=FALSE, message=FALSE}
county.winner <- election.raw %>%
  mutate_at(vars(state, county), tolower) %>%
  mutate(county = gsub(" county| parish", "", county)) %>% # removes county or parish
  mutate(county = gsub(" cty| plt", "", county)) %>%
  mutate(county = gsub(" townships| townshps", "", county)) %>%
  mutate(county = gsub(" township| pt", "", county)) %>%
  mutate(county = gsub("[.]", "", county)) %>% # removes periods
  #mutate(county = gsub("[.]", "", county)) %>% remove hyphen replace with space
  group_by(state, county) %>% # Some states have the same county name. This will differentiate between counties in different states
  mutate(total = sum(total_votes)) %>%
  mutate(pct = total_votes / total) %>%
  top_n(1,pct) %>%
  ungroup()

# head(county.winner)
```



```{r echo=FALSE, error=FALSE, message=FALSE}
state.winner <- election.raw %>%
  mutate_at(vars(state, county), tolower) %>%
  group_by(state, candidate, party) %>%
  summarise(total_state_votes = sum(total_votes)) %>%
  ungroup() %>%
  group_by(state) %>%
  mutate(total_votes_state = sum(total_state_votes)) %>%
  mutate(pct = total_state_votes / total_votes_state) %>%
  top_n(1, pct) %>%
  ungroup() %>%
  select(state, candidate, party, total_state_votes, pct)

# election.raw %>% select(state, county) %>% distinct()

# head(state.winner)
```

# Visualization


```{r echo=FALSE}
electoral_votes_2020 <- data.frame(
  state = c("alabama", "alaska", "arizona", "arkansas", "california", 
            "colorado", "connecticut", "delaware", "district of columbia", 
            "florida", "georgia", "hawaii", "idaho", "illinois", "indiana", 
            "iowa", "kansas", "kentucky", "louisiana", "maine", "maryland", 
            "massachusetts", "michigan", "minnesota", "mississippi", "missouri", 
            "montana", "nebraska", "nevada", "new hampshire", "new jersey", 
            "new mexico", "new york", "north carolina", "north dakota", "ohio", 
            "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", 
            "south dakota", "tennessee", "texas", "utah", "vermont", "virginia", 
            "washington", "west virginia", "wisconsin", "wyoming"),
  
  electoral_votes = c(9, 3, 11, 6, 55, 9, 7, 3, 3, 29, 16, 4, 4, 20, 11, 
                      6, 6, 8, 8, 4, 10, 11, 16, 10, 6, 10, 3, 5, 6, 4, 14, 
                      5, 29, 15, 3, 18, 7, 7, 20, 4, 9, 3, 11, 38, 6, 3, 13, 
                      12, 5, 10, 3),
  
  stringsAsFactors = FALSE
)
```


The R package ggplot2 can be used to draw maps. Consider the following code.

```{r  warning=FALSE}
states <- map_data("state")

ggplot(data = states) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  theme_void() + # removes 
  guides(fill="none")  # color legend is unnecessary and takes too long
```

The variable states contain information to draw white polygons, and fill-colors are determined by region.

**6. (4 pts) Use similar code to above to draw county-level map by creating counties = map_data("county"). Color by county.**

```{r echo=FALSE}
counties <- map_data("county") # head(counties)

ggplot(data = counties) + 
  geom_polygon(aes(x = long, y = lat, fill = region, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  theme_void() +
  guides(fill="none")
```




```{r echo=FALSE}
new_state <- states
# new_state$region <- str_to_title(new_state$region)

new_state <- state.winner %>%
  left_join(new_state, # # state.winner %>% mutate(state = tolower(state))
            by = c("state" = "region"))

#new_state <- new_state %>%
#  left_join(state.winner, # # state.winner %>% mutate(state = tolower(state))
#            by = c("region" = "state"))
# new_state %>% filter(is.na(region))

#new_state <- new_state %>%
#  left_join(electoral_votes_2020,
#            by = c("region" = "state"))
```



```{r echo=FALSE}
ggplot(data = new_state) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  theme_void() +
  guides() + 
  # ggtitle("Presidential Election") +
  theme(plot.title=element_text(hjust=0.5))
```

**8. (6 pts) Color the map of the state of California by the winning candidate for each county.**


```{r echo=FALSE}
counties <- map_data("county")
new_counties <- counties

#new_counties <- new_counties %>%
#  left_join(county.winner,
#            by = c("region" = "state", "subregion" = "county"))

new_counties <- county.winner %>%
  left_join(new_counties,
            by = c("state" = "region", "county" = "subregion"))

#new_winner <- county.winner %>%
#  left_join(counties,
#            by = c("state" = "region", "county" = "subregion"))

```

```{r echo=FALSE}
# new_counties %>% filter(is.na(candidate))

ggplot(data = new_winner) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  theme_void() +
  guides() # + 
  # ggtitle("USA Presidential Election")
```


```{r echo=FALSE}
# new_counties %>% filter(is.na(candidate))
california <- new_counties[which(new_counties$region == 'california'), ]
ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, group = group),
               color = "white") + 
  coord_fixed(1.3) +
  labs(fill = "Candidate") +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.2)) +
  guides() # + # color legend is unnecessary and takes too long
  # ggtitle("California")
```

**9. (4 pts) (Open-ended) Create a visualization of your choice using census data.** Many exit polls noted that demographics played a big role in the election. Use this Washington Post article and this R graph gallery for ideas and inspiration.


```{r echo=FALSE, message=FALSE, warnings=FALSE}
# Heat map
cali_map_og <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, alpha = pct, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  scale_alpha_continuous(range = c(0.4, 1), guide = "none") + # Adjust alpha range for visibility
  theme_minimal() +
  labs(fill = "Candidate", alpha = "Vote %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  theme(plot.title = element_text(hjust=0.2)) +
  ggtitle("Heat Map")

cali_map_og
```


```{r echo=FALSE}
counties <- map_data("county")

# temp <- census %>% mutate(County = as.character(gsub(" County", "", County)))
temp <- census %>% 
  mutate_at(vars(State, County), tolower) %>%
  mutate(County = gsub(" county|  parish", "", County))

temp_new_counties <- new_counties %>%
  left_join(temp, by = c("region" = "State", "subregion" = "County"))

# temp_new_counties <- new_counties

temp_new_counties <- temp_new_counties %>%
  mutate(majority_race = case_when(
    Hispanic > White & Hispanic > Black & Hispanic > Asian ~ "Hispanic",
    White > Hispanic & White > Black & White > Asian ~ "White",
    Black > Hispanic & Black > White & Black > Asian ~ "Black",
    Asian > Hispanic & Asian > White & Asian > Black ~ "Asian",
    TRUE ~ "Other"
  ))
```



```{r echo=FALSE}
california <- temp_new_counties[which(temp_new_counties$region == 'california'), ]

cali_white_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = White, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#D6EAF8", high = "#16537E") + # scale_fill_gradient(low = '#E8E2CC', high = "#004434")
  theme_minimal() +
  labs(fill = "White %") +
  theme(text = element_text(size = 10)) +
  theme_void()# + ggtitle("White Percentage in California Counties")

# Hispanic Percentage in California Counties
cali_hisp_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = Hispanic, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#FADBD8", high = "#D35400") +
  theme_minimal() +
  labs(fill = "Hispanic %") +
  theme(text = element_text(size = 10)) +
  theme_void()

# Asian Percentage in California Counties
cali_asian_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = Asian, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#F5F5F5", high = "#27AE60") +
  theme_minimal() +
  labs(fill = "Asian %") +
  theme(text = element_text(size = 10)) +
  theme_void()


# Black Percentage in California Counties
cali_black_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = Black, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#E8DAEF", high = "#8E44AD") + #scale_fill_gradient(low = '#DDEAD4', high = "#162D4A")
  theme_minimal() +
  labs(fill = "African-American %") +
  theme(text = element_text(size = 10)) +
  theme_void()
```



```{r echo=FALSE}
(cali_white_map + cali_hisp_map) / (cali_asian_map + cali_black_map)  +
  plot_annotation(title = "California Demographics by County",
                  theme = theme(plot.title = element_text(hjust = 0.5, margin = margin(b = 25)))
                  ) &
  theme(plot.margin = margin(15, 15, 15, 15))

```


```{r echo=FALSE}
cali_demog_map <- ggplot(data = california) + 
  geom_polygon(aes(x = long, y = lat, fill = majority_race, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Hispanic" = "#D35400", "White" = "#16537E", "Black" = "#8E44AD", "Asian" = "#27AE60", "Other" = "gray")) +
  scale_alpha_continuous(range = c(0.4, 1), guide = "none") + # Adjust alpha range for visibility
  theme_minimal() +
  labs(fill = "Majority Race") +
  theme(text = element_text(size = 10)) + 
  theme_void() +
  ggtitle("Majority Demographic Group")

cali_map_og+cali_demog_map
```



```{r echo=FALSE}
texas <- temp_new_counties[which(temp_new_counties$region == 'texas'), ]
# candidate is missing for subregion "de witt"
# texas %>% filter(is.na(candidate))

texas_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = candidate, alpha = pct, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Joe Biden" = "blue", "Donald Trump" = "red")) +
  scale_alpha_continuous(range = c(0.4, 1), guide = "none") + # Adjust alpha range for visibility
  theme_minimal() +
  labs(fill = "Candidate", alpha = "Vote %") +
  theme(text = element_text(size = 10)) +
  theme_void() +
  ggtitle("Heat Map of Candidates Texas Counties")

texas_map
```


```{r echo=FALSE}

# White Percentage in Texas Counties
texas_white_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = White, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = '#D6EAF8', high = "#16537E") + # scale_fill_gradient(low = '#E8E2CC', high = "#004434")
  theme_minimal() +
  labs(fill = "White %") +
  theme(text = element_text(size = 10)) +
  theme_void()


# Hispanic Percentage in Texas Counties
texas_hisp_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = Hispanic, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#FADBD8", high = "orange") +
  theme_minimal() +
  labs(fill = "Hispanic %") +
  theme(text = element_text(size = 10)) +
  theme_void()


# Asian Percentage in Texas Counties
texas_asian_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = Asian, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#F5F5F5", high = "#27AE60") +
  theme_minimal() +
  labs(fill = "Asian %") +
  theme(text = element_text(size = 10)) +
  theme_void()

# Black Percentage in Texas Counties
texas_black_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = Black, group = group), color = "white") +
  coord_fixed(1.3) +
  scale_fill_gradient(low = "#E8DAEF", high = "#8E44AD") + #scale_fill_gradient(low = '#DDEAD4', high = "#162D4A")
  theme_minimal() +
  labs(fill = "African-American %") +
  theme(text = element_text(size = 10)) +
  theme_void()

```


```{r echo=FALSE}
(texas_white_map + texas_hisp_map)/ (texas_black_map + texas_asian_map)   +
  plot_annotation(title = "Texas Demographics by County",
                  theme = theme(plot.title = element_text(hjust = 0.5, margin = margin(b = 25)))
                  ) &
  theme(
    plot.margin = margin(2, 2, 2, 2),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```


```{r echo=FALSE}
texas <- temp_new_counties[which(temp_new_counties$region == 'texas'), ]

texas_demog_map <- ggplot(data = texas) + 
  geom_polygon(aes(x = long, y = lat, fill = majority_race, group = group), color = "white") + 
  coord_fixed(1.3) +
  scale_fill_manual(values = c("Hispanic" = "#D35400", "White" = "#16537E", "Black" = "#8E44AD", "Asian" = "#27AE60", "Other" = "gray")) +
  theme_minimal() +
  labs(fill = "Majority Race") +
  theme(text = element_text(size = 10)) + theme_void() +
  ggtitle("Majority Demographic Group")

texas_map + texas_demog_map
```
