---
title: "ML Analysis of the 2020 Presidential Election"
author: "Alexander Sanchez"
output: github_document
---


```{r setup, echo=FALSE, include=FALSE}
# install.packages("arulesViz")
library(knitr)
library(ISLR) 
library(tree)
library(maptree)
library(randomForest) 
library(gbm) 
library(glmnet) 
library(caret)
library(psych)
library(boot)
library(tidyverse)
library(corrplot)
library(FactoMineR)
library(dbplyr)
library(ROCR) 
library(ggridges) 
library(dendextend) 
library(e1071)
library(maps)
library(utils)
library(car)
library(factoextra)
library(Matrix)
library(arules)
library(cluster)
library(arulesViz)
library(ggplot2)
library(dplyr)
library(patchwork)
#update.packages()
# set global chunk options: images will be 8x5 inches
knitr::opts_chunk$set(fig.width=8, fig.height=5)
options(digits = 4)


## indents are for indenting r code as formatted text
## They may need to be adjusted depending on your OS
# if your output looks odd, increase or decrease indent
indent1 = '    '
indent2 = '        '
indent3 = '            '
```




# Instructions and Expectations

Predicting voter behavior is complicated for many reasons despite the tremendous effort in collecting, cleaning, analyzing, and understanding many available datasets. As the midterm elections in the United States is currently being held near the midpoint of the current president’s four-year term of office, it is a good time for us to review the 2020 United States presidential election data! Despite that the 2016 presidential election came as a big surprise to many, Biden’s victory in the 2020 presidential election has been widely predicted (e.g., see the well-known Nate Silver in FiveThirtyEight).


# Data

We will start the analysis with two data sets. The first one is the election data, which is drawn from here. The data contains county-level election results.

The second dataset is the 2017 United States county-level census data, which is available here.

The following code load in these two data sets: `election.raw` and `census.`

```{r message=FALSE, echo=FALSE}
## read data and convert candidate names and party names from string to factor
## we manually remove the variable "won", the indicator of county level winner
## In Problem 5 we will reproduce this variable!
election.raw <- read_csv("../data/raw/candidates_county.csv", col_names = TRUE) %>% 
  mutate(candidate = as.factor(candidate), party = as.factor(party), won = NULL)

## remove the word "County" from the county names
words.to.remove = c("County")
remove.words <- function(str, words.to.remove){
  sapply(str, function(str){
    x <- unlist(strsplit(str, " "))
    x <- x[!x %in% words.to.remove]
    return(paste(x, collapse = " "))
  }, simplify = "array", USE.NAMES = FALSE)
}
election.raw$county <- remove.words(election.raw$county, words.to.remove)


## read census data
census <- read_csv("../data/raw/census_county.csv") 
```


# Data Wrangling

**5. (6 pts) Create data sets county.winner and state.winner by taking the candidate with the highest proportion of votes in both county level and state level. Hint: to create county.winner, start with election.raw, group by state and county, compute total votes, and pct = votes/total as the proportion of votes. Then choose the highest row using top_n (variable state.winner is similar).**


```{r echo=FALSE, error=FALSE, message=FALSE}
county.winner <- election.raw %>%
  mutate_at(vars(state, county), tolower) %>%
  mutate(county = gsub(" county| parish", "", county)) %>% # removes county or parish
  mutate(county = gsub(" cty| plt", "", county)) %>%
  mutate(county = gsub(" townships| townshps", "", county)) %>%
  mutate(county = gsub(" township| pt", "", county)) %>%
  mutate(county = gsub("[.]", "", county)) %>% # removes periods
  group_by(state, county) %>% # Some states have the same county name. This will differentiate between counties in different states
  mutate(total = sum(total_votes)) %>%
  mutate(pct = total_votes / total) %>%
  top_n(1,pct) %>%
  ungroup()

```

```{r echo=FALSE, error=FALSE, message=FALSE}
state.winner <- election.raw %>%
  mutate_at(vars(state, county), tolower) %>%
  group_by(state, candidate, party) %>%
  summarise(total_state_votes = sum(total_votes)) %>%
  ungroup() %>%
  group_by(state) %>%
  mutate(total_votes_state = sum(total_state_votes)) %>%
  mutate(pct = total_state_votes / total_votes_state) %>%
  top_n(1, pct) %>%
  ungroup() %>%
  select(state, candidate, party, total_state_votes, pct)

# election.raw %>% select(state, county) %>% distinct()
```


# Data Cleaning

We start by cleaning the county-level census data through several transformations. Initially, we remove any rows with missing values using `na.omit()` and `filter()`. Then, we convert three attributes (Men, Employed, and VotingAgeCitizen) to percentages by dividing each by TotalPop and multiplying by 100. The code creates a new Minority attribute by combining Hispanic, Black, Native, Asian, and Pacific populations and removes these original variables after creating the Minority column using the `.keep = "unused" argument.` The Minority column is then relocated to appear after the White column for better organization.

The code removes several specified columns that are not needed for the analysis: IncomeErr, IncomePerCap, IncomePerCapErr, Walk, PublicWork, and Construction.

```{r}
census.clean <- census %>%
  na.omit() %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(
    Men = (Men/TotalPop) * 100,
    #Women = (Women/TotalPop) * 100
    Employed = (Employed/TotalPop) * 100,
    VotingAgeCitizen = (VotingAgeCitizen/TotalPop) * 100,
    Minority = Hispanic + Black + Native + Asian + Pacific
    ) %>%
  relocate(Minority, .after = White) %>%
  select(-c(Hispanic, Black, Native, Asian, Pacific, Women)) %>%
  select(-c(IncomeErr, IncomePerCap, IncomePerCapErr, Walk, PublicWork, Construction))

head(census.clean, 5)
```


# Dimensionality reduction

After removing the State and County columns, the code creates a two-column data frame called pc.county containing the first two principal components (PC1 and PC2) from the PCA analysis of the cleaned census data.

```{r echo=FALSE}
temp_census.clean <- census.clean %>%
  select(-c(CountyId, State, County))
```

Regarding centering and scaling: The code uses both centering `(center=TRUE)` and scaling `(scale=TRUE)` before running PCA, which is appropriate because the variables in the census data have different scales. For example, some variables are percentages, while others might be raw counts or dollar values. Without scaling, variables with more significant variances would dominate the principal components, regardless of their importance.

```{r}
pr.out.census <- prcomp(temp_census.clean, scale=TRUE, center = TRUE)
```


```{r echo=FALSE}
# pc.county <- pr.out.census$x[,1:2]
pc.county <- data.frame(PC1 = pr.out.census$x[, 1], PC2 = pr.out.census$x[, 2])
```


The three features with the largest absolute values of the first principal component are:
```{r echo=FALSE}
new_p <- pr.out.census$rotation[,1]
largest_abs_loadings <- sort(abs(new_p), decreasing = TRUE)[1:3]
largest_features <- names(largest_abs_loadings)
```
`r names(largest_abs_loadings)`



```{r message=FALSE, warnings=FALSE, results='hide', echo=FALSE, fig.height = 8, fig.width = 12}
# Extract loadings
loading_df <- data.frame(
  Taxon = colnames(temp_census.clean),  # Replace with the actual dataset column names
  PC1 = pr.out.census$rotation[, 1],      # First principal component
  PC2 = pr.out.census$rotation[, 2]       # Second principal component
)

loading_plot_df <- loading_df %>%
  pivot_longer(
    cols = starts_with("PC"),  # Columns to pivot (e.g., PC1, PC2)
    names_to = "Principal_Component",
    values_to = "Loading"
  )

# Base plot
base <- ggplot(loading_plot_df, aes(x = Loading, y = Taxon, color = Principal_Component)) +
  geom_point() +
  geom_line(aes(group = Taxon)) +  # Group lines by Taxon
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray", linewidth = .6) +  # Horizontal line at zero
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 7),
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14)  
    ) +
  labs(x = "Loading", y = "", color = "Principal Component")

fig7 <- base + facet_wrap(~Principal_Component, ncol = 2)

fig7
```

### PC1 Loading Analysis

Looking at the PC1 loadings, several key pairs of features have opposite signs:

**Key opposite sign pairs:**

- **Poverty/Child Poverty (negative) vs. Income (positive)**: These opposite signs indicate that counties with higher poverty rates tend to have lower incomes, showing a strong negative correlation between economic hardship and affluence.

- **Unemployment (negative) vs. Employed (positive)**: The opposite loadings indicate that counties with high unemployment rates tend to have low employment rates, which is expected given that these are complementary measures.

- **Minority (negative) vs White (positive)**: This indicates that counties with higher minority populations tend to have lower white populations, reflecting demographic composition patterns.

- **Service (negative) vs Professional/Office (positive)**: Counties with more service sector employment tend to have fewer professional and office workers, suggesting different economic structures.

These opposite signs in PC1 indicate that the first principal component captures a socioeconomic divide, where affluent counties (characterized by high income, employment, professional work, and a white population) contrast with economically disadvantaged counties (characterized by high poverty, unemployment, service work, and minority populations). The negative correlations between these feature pairs indicate they tend to move in opposite directions across counties.

### PC2 Loading Analysis

Looking at the PC2 loadings, several key pairs of features have opposite signs:

**Key opposite sign pairs:**

- **Income (negative) vs. White (positive)**: This suggests that in PC2's dimension, areas with higher incomes tend to have lower white populations, indicating PC2 may capture urban diversity patterns where diverse, higher-income areas contrast with predominantly white areas.

- **Transit/OtherTransp (negative) vs. Drive (positive)**: Counties with more public transit and alternative transportation have fewer people driving to work, reflecting urban versus suburban/rural transportation patterns.

- **MeanCommute (negative) vs. WorkAtHome (positive)**: Areas with longer average commutes tend to have fewer people working from home, suggesting PC2 captures commuting versus remote work patterns.

- **Service (negative) vs. Professional/Office (positive)**: This indicates different employment structures, where service-oriented economies contrast with professional or office-based economies.

PC2 captures an urban-suburban/rural divide where dense, transit-oriented areas with diverse populations and longer commutes contrast with regions characterized by car dependency, remote work, and professional employment. The opposite signs reveal that PC2 identifies counties along a spectrum of urbanization and work/transportation patterns.


```{r echo=FALSE}
##### **Combine PCA results with county and state**
# Extract the principal components
pca_data <- as.data.frame(pr.out.census$x)

# Add the county id, county, and state information back to the PCA results
pca_data$CountyId <- census.clean$CountyId
pca_data$County <- census.clean$County
pca_data$State <- census.clean$State
pca_data <- pca_data %>%
  select(CountyId, County, State, everything())

# head(pca_data)
```


## Determine the number of minimum number of PCs needed to capture 90% of the variance for the analysis

The code calculates the proportion of variance explained (PVE) for each principal component by:
1. Extracting the standard deviations from the principal component analysis
2. Calculating the variance by squaring the standard deviations
3. Computing the proportion of variance by dividing each component's variance by the total variance

```{r}
x <- pr.out.census$sdev

pr.var <- x^2

pve <- pr.var/sum(pr.var)
```

```{r echo=FALSE, message=FALSE, results='hide'}
# We want the cumulative at each index. The index is the PC. So index 1 is PC 1 
# and so on. We then want to know at which index is when it is greater than or equal to
# 90% ergo which(cumsum(pve) >= .9). Once we know which indexes fit this condition,
# we want to know the first index to fit this condition ergo min(...). 
# Therefore: 
# min(which(cumsum(pve) >= .9))
```
Computing `min(which(cumsum(pve) >= .9))`, we need about `r min(which(cumsum(pve) >= .9))` PCs in order to explain 90% of the total variation in the data. Thus, we need to retain the first `r min(which(cumsum(pve) >= .9))` components to explain at least 90% of the variability in the original data.


The cumulative proportion of variance plot visually demonstrates how the explained variance accumulates as more principal components are included, allowing us to see the incremental contribution of each additional component to the total variance explained.

```{r pve-plot, echo=FALSE, fig.width = 12}
# Creating PVE dataframe
pca_df <- data.frame(
  component = 1:length(pve),
  proportion_variance = pve
)

# Individual PVE plot
p1 <- ggplot(pca_df, aes(x=component, y=proportion_variance)) + 
  geom_point(size=2) +
  geom_line(linewidth=1) + 
  labs(title = "Individual PVE",
       x = "Principal Component", 
       y = "Proportion of Variance Explained") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(1, dim(pca_df)[1], 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Cumulative PVE plot
p2 <- ggplot(pca_df, aes(x=component, y=cumsum(proportion_variance))) + 
  geom_point(size=2) +
  geom_line(linewidth=1) + 
  labs(title = "Cumulative PVE",
       x = "Principal Component", 
       y = "Cumulative Proportion") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = seq(1, dim(pca_df)[1], 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


# Combine plots
p1 + p2 + 
  plot_annotation(
    title = "PCA Variance Analysis", 
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5))
    )
```

### PCA Justification and Interpretation

**1. Is focusing on PC1 and PC2 justified?**

- PC1 and PC2 together account for roughly 38-40% of the total variance, which is low compared to the usual 70-80% threshold for good representation. However, the sharp elbow after PC2 in the individual PVE plot and the significant drop from about 14% to 7% suggest these two components capture the most meaningful patterns in the data.

**2. What overall patterns do PC1 and PC2 capture?**

- PC1 (~25% variance) reflects a socioeconomic divide where affluent counties (high income, employment, professional work, white population) contrast with disadvantaged counties (high poverty, unemployment, minority populations). PC2 (~13-14% variance) indicates an urban-suburban/rural divide characterized by transportation patterns (transit vs. driving), work arrangements (commuting vs. remote), and population density differences.


# Clustering

```{r echo=FALSE}
census.clean <- census %>%
  na.omit() %>%
  filter(if_any(everything(), ~ !is.na(.))) %>%
  mutate(
    Men = (Men/TotalPop) * 100,
    #Women = (Women/TotalPop) * 100
    Employed = (Employed/TotalPop) * 100,
    VotingAgeCitizen = (VotingAgeCitizen/TotalPop) * 100,
    Minority = Hispanic + Black + Native + Asian + Pacific
    ) %>%
  relocate(Minority, .after = White) %>%
  select(-c(Hispanic, Black, Native, Asian, Pacific, Women)) %>%
  select(-c(IncomeErr, IncomePerCap, IncomePerCapErr, Walk, PublicWork, Construction))
```

### Centering and Scaling
Same approach as dimensionality reduction.
```{r}
scar <- scale(census.clean[,c(-1:-3)], center=TRUE, scale=TRUE)
census.clean.dist <- dist(scar)
set.seed(10)
census.clean.hclust <- hclust(census.clean.dist)
```

### First Clustering (Original Features):

```{r echo=FALSE}
set.seed(10)
# Get cluster
clus <- cutree(census.clean.hclust, 10)
table_data <- as.data.frame(table(clus))
```

When performing hierarchical clustering using the scaled original features with complete linkage and cutting the tree into 10 clusters, the distribution of observations across clusters is quite uneven. Cluster 1 contains the most observations, with `r table(clus)["1"]` counties, while several other clusters have very few counties. For instance, clusters 6 and 10 have only `r table(clus)["6"]` and `r table(clus)["10"]` counties, respectively, indicating that the clustering algorithm identified a few very distinct groups among the counties.

```{r echo=FALSE}
# table(clus)
ggplot(table_data, aes(x = factor(clus), y = Freq)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = Freq), vjust = -0.3, size = 3) +
  labs(title = "County Distribution",
       x = "Cluster",
       y = "Number of Counties") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) 
```



### Second Clustering (First Two Principal Components):

```{r echo=FALSE}
census_transform <- as.data.frame(pc.county)

census_transform.dist <- dist(pc.county)

set.seed(10)
census_transform.hclust <- hclust(census_transform.dist)
clus_new <- cutree(census_transform.hclust, 10)
table_data <- as.data.frame(table(clus_new))
```

The cluster distribution changes significantly after re-running the hierarchical clustering using the first two principal components from pc.county. Cluster 1 contains `r table(clus_new)["1"]` counties, cluster 2 has `r table(clus_new)["2"]` counties, and Cluster 4 has `r table(clus_new)["4"]` counties. The remaining clusters have fewer counties, but the distribution differs from the first clustering.

```{r echo=FALSE}
ggplot(table_data, aes(x = factor(clus_new), y = Freq)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = Freq), vjust = -0.3, size = 3) +
  labs(title = "County Distribution",
       x = "Cluster",
       y = "Number of Counties") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 
```

What can we take out of this

* Both clustering approaches result in highly imbalanced cluster sizes, suggesting some very distinct groups of counties.
* The uneven cluster sizes in both methods suggest that a few very distinct county groups stand out from the majority, which could represent counties with unique demographic or economic characteristics.


# Classification

We exclude the predictor party from election.cl because the classification task aims to predict a county's winner using census information. Including party as a predictor would create data leakage, as it directly informs us about the candidate's party affiliation rather than the demographic or socioeconomic factors of the county.

The party variable is tied to the outcome (the winner), and its inclusion would undermine the model's ability to generalize by basing predictions on the relationship between party and candidate rather than the census data. We remove the party variable from the dataset to ensure the model learns from relevant predictors.

```{r}
# we move all state and county names into lower-case
tmpwinner <- county.winner %>% ungroup %>%
  mutate_at(vars(state, county), tolower)

# we move all state and county names into lower-case
# we further remove suffixes of "county" and "parish"
tmpcensus <- census.clean %>% mutate_at(vars(State, County), tolower) %>%
  mutate(County = gsub(" county|  parish", "", County))

# we join the two datasets
election.cl <- tmpwinner %>%
  left_join(tmpcensus, by = c("state"="State", "county"="County")) %>%
  na.omit

# drop levels of county winners if you haven't done so in previous parts
election.cl$candidate <- droplevels(election.cl$candidate)

## save meta information
election.meta <- election.cl %>% select(c(county, party, CountyId, state, total_votes, pct, total))

## save predictors and class labels
election.cl = election.cl %>% select(-c(county, party, CountyId, state, total_votes, pct, total))
```


## Train/Test Split

Using the following code, partition data into 80% training and 20% testing:

```{r}
set.seed(10)
n <- nrow(election.cl)
idx.tr <- sample.int(n, 0.8*n)
election.tr <- election.cl[idx.tr, ]
election.te <- election.cl[-idx.tr, ]
```

Using the following code to define 10 cross-validation folds:

```{r}
set.seed(10)
nfold <- 10
folds <- sample(cut(1:nrow(election.tr), breaks=nfold, labels=FALSE))
```

Using the following error rate function. And the object records is used to record the classification performance of each method in the subsequent problems.

```{r}
calc_error_rate = function(predicted.value, true.value){
  return(mean(true.value!=predicted.value))
}
records = matrix(NA, nrow=6, ncol=2)
colnames(records) = c("train.error","test.error")
rownames(records) = c("Decision Tree","Logistic Regression","Lasso Logistic Regression","Random Forest","Neural Network", "Neural Network Optimized")
```


## Decision Tree Analysis

In this analysis, we'll build a decision tree classifier to predict county-level voting patterns in the 2020 election. Decision trees are especially useful for understanding complex data relationships because they create interpretable rules that reflect human decision-making processes. We'll begin by training a full decision tree on our training data, then use cross-validation to find the best tree size that minimizes misclassification error while preventing overfitting. After pruning the tree to this optimal size, we'll visualize both the original and pruned trees to see how the model makes predictions. Finally, we'll evaluate the model's performance on both training and test sets and interpret the decision rules to gain insights into American voting behavior and the factors influencing political preferences at the county level.


### Building the Initial Tree

First, we train a decision tree using all available predictors:
```{r echo=FALSE}
tree.election <- tree(candidate~.,data = election.tr)
```

```{r echo=FALSE}
draw.tree(tree.election, nodeinfo=TRUE, cex = 0.5)
title("Single Tree Built on Training Set")
```

```{r echo=FALSE}
set.seed(10)
cv.election <- cv.tree(tree.election, FUN = prune.misclass, rand = folds)

# Find all sizes with minimum error
min_error <- min(cv.election$dev)
best_sizes <- cv.election$size[cv.election$dev == min_error]

# The largest size among tied minimums
best_size_max <- max(best_sizes)

# The smallest size among tied minimums
best_size_min <- min(best_sizes)
```


### Cross-Validation and Pruning

We use cross-validation to find the optimal tree size that minimizes misclassification error:

Cross-validation suggests the optimal tree size is `r best_size_min nodes`.

```{r echo=FALSE}
cv_data <- data.frame(
  size = cv.election$size,
  error = cv.election$dev
)

# Create the plot
ggplot(cv_data, aes(x = size, y = error)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "Cross-Validation Error vs Tree Size",
    x = "Tree Size",
    y = "Cross-Validation Error"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(breaks = unique(cv_data$size)) +  # All tree sizes on x-axis
  geom_point(
    data = cv_data[cv_data$error == min(cv_data$error), ],
    aes(x = size, y = error),
    color = "red",
    size = 4 
  )  # Highlight minimum error point
```


### Pruned Tree

We prune the tree to the optimal size and visualize the result:
```{r echo=FALSE}
# Prune Tree
pt.cv <- prune.misclass(tree.election, best = best_size_min)
```

```{r echo=FALSE}
# Plot Pruned Tree
draw.tree(pt.cv, nodeinfo=TRUE, cex = 0.5)
title(paste("Pruned Tree of Size ", best_size_min))
```


## Decision Tree Analysis: A Story of American Voting Patterns

This decision tree provides fascinating insights into county-level voting patterns in the 2020 election, offering a clear overview of the different types of American communities.

### The Primary Divide: Transit Usage

The tree's **root split on Transit > 1.05%** immediately divides America into two distinct worlds.

- **Left branch (Transit ≤ 1.05%)**: Rural and suburban areas in America with limited public transit.
- **Right branch (Transit > 1.05%)**: More urban areas with public transit.

This indicates that **urbanization level** (measured by transit usage) was the most predictive factor for voting behavior.

### Rural/Suburban America Story (Left Branch)

In low-transit areas, the tree tells a nuanced story:

**1. The "White Rural" Pattern (Node 4)**

- Counties with **more than 48.9% White population** → **Donald Trump (1,823 counties, 91.7% accuracy)**
- This depicts traditional rural America - mostly white, low transit usage.
- Massive Trump stronghold, the largest single group.

**2. The "Diverse Rural" Pattern**

- Counties with **48.9% or less White population** further divided by unemployment:
  - **High unemployment (>7.85%)** → **Joe Biden (Node 3, 102 counties)**
  - **Low unemployment (≤7.85%)** → Further subdivisions on voting-age citizenship

**3. The "Mixed Communities" Pattern (Nodes 1 & 2)**

- In various rural areas with low unemployment, **voting-age citizenship** becomes crucial.
- **Lower citizenship rates (≤70.3%)** → **Trump (Node 1, 48 counties)**
- **Higher citizenship rates (>70.3%)** → **Biden (Node 2, 25 counties)**

### Urban America Story (Right Branch)

In higher-transit areas, the pattern is simpler but revealing:

- **Small-medium urban areas (population ≤136K)** → **Trump (Node 5, 202 counties)**
- **Large urban areas (population >136K)** → **Biden (Node 6, 208 counties)**

### Key Voting Behavior Insights

1. **Geographic Polarization**: The main divide is between urban and rural areas, as indicated by transit usage.

2. **Racial Demographics Matter**: In rural areas, racial makeup is the second most important factor.

3. **Economic hardship influences voting for Democrats**: High unemployment in diverse rural areas favors Biden.

4. **Urban Size Threshold**: There is a critical population size (~136K) at which urban areas switch from Republican to Democratic.

5. **Immigration Patterns**: In mixed rural communities, areas with more recent immigrants (those with lower voting-age citizenship) tend to support Trump, possibly reflecting concerns about economic competition.

### The Broader Narrative

This tree captures the **"Two Americas"** phenomenon:

- **Rural and small-town America**: Mainly Republican, with exceptions in economically distressed and diverse regions.
- **Urban America**: Tends to be Democratic in major cities, but smaller urban areas can still lean Republican
- **The complexity of modern politics**: Even within these broad categories, factors like economic conditions, demographics, and immigration patterns add important nuances.



```{r echo=FALSE}
# Predict on test set
pred.train.tree = predict(pt.cv, election.tr, type="class")
pred.test.tree = predict(pt.cv, election.te, type="class")

# Obtain confusion matrix
err.pt.cv <- table(pred.test.tree, election.te$candidate) # table(pred.train.tree, election.tr$candidate)
err.pt.cv
```

```{r echo=FALSE}
a <- calc_error_rate(pred.train.tree, election.tr$candidate)
b <- calc_error_rate(pred.test.tree, election.te$candidate)
records[1,] <- c(a,b)
```

## Logistic Regression Analysis

We'll now develop a logistic regression model to predict county-level voting outcomes and compare its insights with our decision tree analysis. Unlike decision trees that generate interpretable rules through splits, logistic regression estimates the outcome's probability by modeling the relationship between predictors and the log-odds of voting for a specific candidate. This method allows us to measure the individual impact of each variable while accounting for all others.

Our analysis will involve fitting the logistic regression model to our training data, evaluating its performance on both training and test sets, and identifying which variables significantly influence voting patterns. We'll examine both statistical significance (p-values) and practical significance (odds ratios) to understand which factors have meaningful real-world impact. Finally, we'll compare these results with our decision tree findings to see whether different modeling approaches reveal consistent patterns in American voting behavior or highlight different aspects of the electoral landscape.


```{r logistic-reg-fitting, echo=FALSE}
# Fitting model using logistic regression
glm.obj.election <- glm(candidate~.,data = election.tr, family = binomial)
```

**Coefficient Interpretation:**

```{r echo=FALSE}
(exp(tail(sort(glm.obj.election$coefficients[2:24]), 5)) - 1)*100
```

Let's focus on VotingAgeCitizen as a specific example. A 1 percentage point increase in VotingAgeCitizen is linked to a 22.70% rise in the odds of a county voting for Joe Biden, holding all other variables constant. This indicates that counties with higher proportions of voting-age citizens tend to favor Democratic candidates.

## Significant Variables

#### P-Value Approach (p < 0.05)

```{r log-reg-sum-pval, echo=FALSE}
# Traditional statistical significance approach
model_summary <- summary(glm.obj.election)
coef_table <- model_summary$coefficients

# Variables with p < 0.05
significant_vars <- rownames(coef_table)[coef_table[,4] < 0.05]
significant_vars <- significant_vars[significant_vars != "(Intercept)"]  # Remove intercept

#print(coef_table[coef_table[,4] < 0.05, ])
```

Using traditional $p < 0.05$ criteria, the significant variables are `r significant_vars`. Something we need to consider is our large sample size (n = 2,408). With our large sample size (n = 2,408), nearly all non-zero coefficients will appear statistically significant, making this criterion less meaningful for practical variable selection.

### Odds Ratios (>20% change in odds)

**Why Odds Ratios Over P-Values**: While p-values indicate whether an effect is statistically significant, they do not show the **practical significance** or **magnitude** of the effect. In large datasets like ours, even tiny, practically meaningless effects become statistically significant. Odds ratios display the **size** of the effect, helping us determine which variables have meaningful real-world impacts on voting behavior. 

**Understanding Odds Ratios:** An odds ratio indicates the multiplicative change in odds with a one-unit increase in the predictor. For example, VotingAgeCitizen = 1.2270 means that for each 1 percentage point rise in voting-age citizens, the odds of voting for Biden increase by 22.70%, assuming all other variables stay the same. 


```{r echo=FALSE}
# Get odds ratios for all coefficients
odds_ratios <- exp(glm.obj.election$coefficients[-1]) # Remove intercept

# Focus on meaningful magnitudes - set your own thresholds
meaningful_effects <- odds_ratios[abs(log(odds_ratios)) > log(1.2)]  # > 20% change in odds
print(meaningful_effects)
```

**Important Limitations**: These interpretations assume a ceteris paribus (all other variables held constant) approach, which rarely occurs in real life. Additionally, logistic regression assumes linear relationships between predictors and the log-odds and that observations are independent. The model does not account for complex interactions between variables that might exist in actual voting behavior. 


### Comparison with Decision Tree Analysis 

**Similarities:**

- Both methods identify **White, VotingAgeCitizen, and Unemployment** as key predictors. 
- Both highlight the significance of **demographic composition** and **economic conditions** in voting behavior. 
- Both demonstrate that **higher unemployment** generally benefits Democratic candidates. 

**Key Differences:** 

- **Geographic vs. Occupational Focus**: The decision tree emphasizes **Transit** (urbanization) as the main split, while logistic regression highlights **occupational categories** (Professional, Service) as the key predictors. 
- **Variable Selection**: Decision tree uses **TotalPop** for urban classification, but logistic regression shows **Service** and **Professional** occupations as more influential. 
- **Interpretability**: The decision tree provides clear geographic narratives (urban vs. rural), while logistic regression uncovers more detailed occupational and demographic effects. 

**General Analysis:** Both methods agree on the crucial role of demographic and economic factors in voting behavior, but they highlight different aspects. The decision tree's geographic focus matches the "Two Americas" narrative, while logistic regression's focus on occupational factors suggests that **job type** and **economic role** might be even more predictive than simple urban-rural geography. This complementary view broadens our understanding of American election patterns.


```{r echo=FALSE}
set.seed(10)
prob.training <- predict(glm.obj.election, election.tr, type="response")
prob.test <- predict(glm.obj.election, election.te, type="response")

# Convert probabilities to binary predictions
majority_rule <- 0.5
pred_training <- ifelse(prob.training > majority_rule, "Joe Biden", "Donald Trump")
pred_test <- ifelse(prob.test > majority_rule, "Joe Biden", "Donald Trump")
```

```{r echo=FALSE}
# Training set confusion matrix and error calculation
training_pred_table <- table(Predicted = pred_training, True = election.tr$candidate)
train_accuracy <- sum(diag(training_pred_table)) / nrow(election.tr)

# Test set confusion matrix and error calculation
test_pred_table <- table(Predicted = pred_test, True = election.te$candidate)
test_accuracy <- sum(diag(test_pred_table)) / nrow(election.te)


c <- calc_error_rate(pred_training, election.tr$candidate)
d <- calc_error_rate(pred_test, election.te$candidate)
records[2,] <- c(c,d)
```


**17. You may notice that you get a warning glm.fit: fitted probabilities numerically 0 or 1 occurred.** As we discussed in class, this is an indication that we have perfect separation (some linear combination of variables perfectly predicts the winner). This is usually a sign that we are overfitting. One way to control overfitting in logistic regression is through regularization.

(3 pts) Use the cv.glmnet function from the glmnet library to run a 10-fold cross validation and select the best regularization parameter for the logistic regression with LASSO penalty. Set lambda = seq(1, 50) * 1e-4 in cv.glmnet() function to set pre-defined candidate values for the tuning parameter λ.

```{r LASSO-train-test-split, echo=FALSE}
# Train-Test split
set.seed(10)
n <- nrow(election.cl)
dat <- model.matrix(candidate~., election.cl)[, -1]  # Remove intercept
idx.tr <- sample.int(n, 0.8*n)

x.train <- dat[idx.tr, ] 
y.train <- election.cl$candidate[idx.tr]

x.test <- dat[-idx.tr, ] 
y.test <- election.cl$candidate[-idx.tr]
```


```{r LASSO-cv-regularization, echo=FALSE}
set.seed(10)

# Tuning a LASSO logistic regression model by finding the optimal regularization
cv.out.lasso <- cv.glmnet(x.train, y.train, 
                          nfolds = 10, alpha = 1, 
                          lambda = (1e-4) * seq(1, 50),
                          family = "binomial")

bestlam <- cv.out.lasso$lambda.min
```

(1 pts) What is the optimal value of λ in cross validation? (1 pts) What are the non-zero coefficients in the LASSO regression for the optimal value of λ? (1 pts) How do they compare to the unpenalized logistic regression? (1 pts) Comment on the comparison. (1 pts) Save training and test errors to the records variable.

The optimal value of $\lambda$ in cross validation is `r bestlam`.

Below are the non-zero coefficients in the LASSO regression for the optimal value of λ:
```{r LASSO-fitting, echo=FALSE}
lasso.model <- glmnet(x.train, y.train, alpha=1, family = "binomial")
pred <- predict(lasso.model,type="coefficients",s=bestlam)[1:24,]
```

```{r extract-LASSO-features, echo=FALSE}
dense_coefficients <- as.matrix(coef(lasso.model, s = bestlam))
non_zero_coefficients <- dense_coefficients[dense_coefficients != 0, ]
print(non_zero_coefficients)
```

```{r LASSO-predict, echo=FALSE}
set.seed(10)

# Predict probabilities for training and test datasets
prob.training <- predict(lasso.model, type = "response", s = bestlam, newx = x.train)
prob.test <- predict(lasso.model, type = "response", s = bestlam, newx = x.test)

# Generate predictions using majority rule
majority_rule <- 0.5
pred_training <- ifelse(prob.training > majority_rule, "Joe Biden", "Donald Trump")
pred_test <- ifelse(prob.test > majority_rule, "Joe Biden", "Donald Trump")
```


```{r LASSO-evaluation, echo=FALSE}
# Training set confusion matrix and error calculation
training_pred_table <- table(Predicted = pred_training, True = election.tr$candidate)
percent_train <- sum(diag(training_pred_table)) / nrow(election.tr) * 100

# Test set confusion matrix and error calculation
test_pred_table <- table(Predicted = pred_test, True = election.te$candidate)
percent_test <- sum(diag(test_pred_table)) / nrow(election.te) * 100

# Error Rate Calculations
e <- calc_error_rate(as.factor(pred_training), y.train)
f <- calc_error_rate(as.factor(pred_test), y.test)
records[3,] <- c(e,f)
```

Unpenalized logistic regression: When taking their absolute values, Women and Income had low coefficients compared to the rest of the variables. Excluding these two, the other variables were within range of each other.

LASSO regression: Minority, Income, WorkAtHome, and MeanCommute had zero coefficients.

It does seem Income is not an important metric. White has been consistently shown as being an important variable.


**18. (6 pts) Compute ROC curves for the decision tree, logistic regression and LASSO logistic regression using predictions on the test data.** Display them on the same plot. (2 pts) Based on your classification results, discuss the pros and cons of the various methods. (2 pts) Are the different classifiers more appropriate for answering different kinds of questions about the election?

```{r model-prediction-set, echo=FALSE}
## Get predicted probabilities from each model for ROC analysis

# Decision Tree - get the probabilities for "Joe Biden"
pred_tree <- predict(pt.cv, election.te, type="vector")[, 2]

# Logistic Regression - get predicted probabilities
pred_log <- predict(glm.obj.election, election.te, type="response") # returns predicted probabilities (between 0 and 1)

# LASSO Logistic Regression - get predicted probabilities using optimal lambda
pred_lasso <- predict(lasso.model, type="response", s=bestlam, newx = x.test)
``` 


```{r roc-analysis-setup, echo=FALSE}
## Create prediction objects for ROC analysis (combines predictions with true labels)

# Decision Tree
roc_pred_tree <-
   prediction(
     predictions = pred_tree,
     labels = election.te$candidate
)

# Logistic Regression
roc_pred_log <-
   prediction(
     predictions = pred_log,
     labels = election.te$candidate
)

# LASSO
roc_pred_lasso <-
   prediction(
     predictions = pred_lasso,
     labels = y.test
)

## Calculate ROC curve coordinates (TPR vs FPR) for each model

# Decision Tree
roc_perf_tree <- performance(roc_pred_tree, measure = "tpr", x.measure = "fpr")

# Logistic Regression
roc_perf_log <- performance(roc_pred_log, measure = "tpr", x.measure = "fpr")

# LASSO
roc_perf_lasso <- performance(roc_pred_lasso, measure = "tpr", x.measure = "fpr")
```


```{r roc-data-preparation, echo=FALSE}
# Combine ROC curve data from all three models into a single data frame for plotting
roc_data <- data.frame(
  # Extract False Positive Rates from all models
  FPR = c(
    unlist(roc_perf_tree@x.values), # Decision Tree
    unlist(roc_perf_log@x.values), # Logistic Regression
    unlist(roc_perf_lasso@x.values)), # LASSO
  
  # Extract True Positive Rates from all models
  TPR = c(
    unlist(roc_perf_tree@y.values), # Decision Tree
    unlist(roc_perf_log@y.values), # Logistic Regression
    unlist(roc_perf_lasso@y.values)), # LASSO
  
  # Create model labels with correct lengths for each model's data points
  Model = factor(rep(c("Tree", "Log", "Lasso"), times = c(length(roc_perf_tree@x.values[[1]]),
                                                          length(roc_perf_log@x.values[[1]]),
                                                          length(roc_perf_lasso@x.values[[1]])
                                                          )
                     )
                 )
)
```

```{r roc-curve-plot, echo=FALSE}
# Create ROC curve comparison plot
ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +  # Plot ROC curves for each model
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +  # Add diagonal reference line (random classifier)
  scale_color_manual(values = c("green", "blue", "red")) +  # Set custom colors for each model
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", # (FPR)
       y = "True Positive Rate", # (TPR)
       color = "Model",
       linetype = "Model") +
  theme_classic() +  # Use clean classic theme
  theme(plot.title = element_text(hjust = 0.5)) +  # Center the title
  theme(legend.position = "bottom")  # Move legend to bottom
```


**Answer:**

**Answer: **

# Taking it further

This part will be worth up to a 20% of your final project grade!

**19. (9 pts) Explore additional classification methods.** Consider applying additional two classification methods from KNN, LDA, QDA, SVM, random forest, boosting, neural networks etc. (You may research and use methods beyond those covered in this course). How do these compare to the tree method, logistic regression, and the lasso logistic regression?


```{r echo=FALSE}
set.seed(10)
n <- nrow(election.cl)
idx.tr <- sample.int(n, 0.8*n)
election.tr <- election.cl[idx.tr, ]
election.te <- election.cl[-idx.tr, ]
```

### Random Forest

```{r echo=FALSE}
# Load required libraries
library(doParallel)
library(parallelly)
```


```{r random-forest-fitting, echo=FALSE, results='hide', error=FALSE, message=FALSE}
# Clean up first
try(stopCluster(cl), silent = TRUE)
registerDoSEQ()

# Setup parallel processing
num_cores <- availableCores(omit = 1)
cl <- makeCluster(num_cores)
registerDoParallel(cl)

ctrl <- trainControl(
  method = "cv", 
  number = 10,
  allowParallel = TRUE,  # ENABLE parallel processing
  verboseIter = TRUE     # Show progress
)

rf_grid <- expand.grid(mtry = c(2, 4, 6, 8, 10))

set.seed(10)
rf_model <- train(candidate~., 
                  data = election.tr,
                  method = "rf",
                  trControl = ctrl,
                  tuneGrid = rf_grid,
                  importance = TRUE,
                  ntree = 500)

stopCluster(cl)
registerDoSEQ()
```

```{r random-forest-predict, echo=FALSE}
# Make predictions on training and test set
pred_training <- predict(rf_model, newdata = election.tr)
pred_test <- predict(rf_model, newdata = election.te)

# Storing Error Rates
a <- calc_error_rate(as.factor(pred_training), election.tr$candidate)
b <- calc_error_rate(as.factor(pred_test), election.te$candidate)
records[4,] <- c(a,b)
```

```{r random-forest-roc-analysis-setup, echo=FALSE}
# Create prediction objects for ROC analysis (combines predictions with true labels)
pred_randfor <- predict(rf_model, newdata = election.te, type = "prob")[, 2]

# Create prediction object for ROC analysis (combines predictions with true labels)
roc_pred_randfor <-
   prediction(
     predictions = pred_randfor,
     labels = election.te$candidate
)

# Calculate ROC curve coordinates (TPR vs FPR)
roc_perf_randfor <- performance(roc_pred_randfor, measure = "tpr", x.measure = "fpr")
```


### Neural Net

```{r echo=FALSE}
set.seed(10)
n <- nrow(election.cl)
idx.tr <- sample.int(n, 0.8*n)
election.tr <- election.cl[idx.tr, ]
election.te <- election.cl[-idx.tr, ]
```

```{r echo=FALSE}
# Load required libraries
library(nnet)
library(caret)  # For data preprocessing
library(parallel)
library(doParallel)
```

```{r neural-net-fitting, echo=FALSE, results='hide', error=FALSE, message=FALSE}
# Clean up any parallel processing connections first
try(stopCluster(cl), silent = TRUE)
registerDoSEQ()

# Scale our predictors for neural network
# Create preprocessing object using training data
preprocess_params <- preProcess(election.tr[,-1], method = c("center", "scale"))

# Apply preprocessing to both training and test sets
X_train_scaled <- predict(preprocess_params, election.tr[,-1])
X_test_scaled <- predict(preprocess_params, election.te[,-1])

# Add the response variable back
train_data_scaled <- data.frame(
  candidate = election.tr$candidate,
  X_train_scaled
)
test_data_scaled <- data.frame(
  candidate = election.te$candidate,
  X_test_scaled
)

# Set seed for reproducibility
set.seed(10)

# Train neural network
nn_model <- nnet(
  candidate ~ ., 
  data = train_data_scaled,
  size = 5,        # number of hidden nodes
  decay = 0.1,     # weight decay for regularization
  maxit = 1000,    # maximum iterations
  trace = FALSE    # suppress training output
)
```


```{r neural-net-predict, echo=FALSE}
# Make predictions on training and test set
train_pred_prob <- predict(nn_model, train_data_scaled, type = "raw")
test_pred_prob <- predict(nn_model, test_data_scaled, type = "raw")

# Convert probabilities to class predictions
train_pred <- ifelse(train_pred_prob > 0.5, "Joe Biden", "Donald Trump")
test_pred <- ifelse(test_pred_prob > 0.5, "Joe Biden", "Donald Trump")

# Storing Error Rates
a <- calc_error_rate(as.factor(train_pred), election.tr$candidate)
b <- calc_error_rate(as.factor(test_pred), election.te$candidate)
records[5,] <- c(a,b)
```

```{r neural-net-roc-analysis-setup, echo=FALSE}
# Create prediction objects for ROC analysis (combines predictions with true labels)
roc_pred_nn <- prediction(
    predictions = test_pred_prob,
    labels = election.te$candidate #as.factor(election.te$candidate, levels = c("Donald Trump", "Joe Biden"))
)

# Calculate ROC curve coordinates (TPR vs FPR)
roc_perf_nn <- performance(roc_pred_nn, measure = "tpr", x.measure = "fpr")
```

### Neural Net Optimized

```{r nn-opt-fitting, echo=FALSE, results='hide', error=FALSE, warning=FALSE, message=FALSE}
# Clean up any parallel processing connections first
try(stopCluster(cl), silent = TRUE)
registerDoSEQ()

# Modify factor levels to be valid R names
election.tr$candidate <- factor(election.tr$candidate,
                              levels = c("Donald Trump", "Joe Biden"),
                              labels = c("Donald_Trump", "Joe_Biden"))
election.te$candidate <- factor(election.te$candidate,
                              levels = c("Donald Trump", "Joe Biden"),
                              labels = c("Donald_Trump", "Joe_Biden"))

# Setup parallel processing
num_cores <- availableCores(omit = 1)  # Leave one core free
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Set seed
set.seed(10)

# Create training control object
ctrl <- trainControl(
  method = "cv",
  number = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  verboseIter = TRUE,
  allowParallel = TRUE,
  seeds= set.seed(10)
)

# Define grid of hyperparameters
grid <- expand.grid(
  size = c(3, 5, 7, 10, 13),
  decay = c(0.001, 0.01, 0.1, 0.3, 0.5)
)


# Train model with grid search
nnet_optimal <- train(
  x = X_train_scaled,
  y = election.tr$candidate,
  method = "nnet",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = grid,
  trace = FALSE,
  maxit = 1000,
  linout = FALSE,
  MaxNWts = 1000
)

# Stop parallel processing
stopCluster(cl)
```

```{r nn-opt-predict, echo=FALSE}
# Make predictions with optimal model
train_pred_prob <- predict(nnet_optimal, X_train_scaled, type = "prob")[,"Joe_Biden"]
test_pred_prob <- predict(nnet_optimal, X_test_scaled, type = "prob")[,"Joe_Biden"]

# Calculate error rates
train_pred <- ifelse(train_pred_prob > 0.5, "Joe_Biden", "Donald_Trump")
test_pred <- ifelse(test_pred_prob > 0.5, "Joe_Biden", "Donald_Trump")

# Storing Error Rates
a <- calc_error_rate(as.factor(train_pred), election.tr$candidate)
b <- calc_error_rate(as.factor(test_pred), election.te$candidate)
records[6,] <- c(a,b)
```

```{r nn-opt-plot, echo=FALSE}
# Plot tuning results
plot(nnet_optimal)
```


```{r nn-opt-roc-analysis-setup, echo=FALSE}

roc_pred_opt <- prediction(
    predictions = test_pred_prob,
    labels = factor(election.te$candidate, levels = c("Donald_Trump", "Joe_Biden"))
)

# Calculate ROC curve coordinates (TPR vs FPR)
roc_perf_opt <- performance(roc_pred_opt, measure = "tpr", x.measure = "fpr")
```


```{r echo=FALSE}
roc_data <- data.frame(
  FPR = c(
    unlist(roc_perf_randfor@x.values), # Random Forest
    unlist(roc_perf_nn@x.values), # Neural Net
    unlist(roc_perf_opt@x.values) # Neural Net Optimized
    ),
  TPR = c(
    unlist(roc_perf_randfor@y.values), # Random Forest
    unlist(roc_perf_nn@y.values), # Neural Net
    unlist(roc_perf_opt@y.values) # Neural Net Optimized
    ),
  Model = factor(rep(c("Random Forest", "Neural Net", "Neural Net Optimized"), 
                     times = c(length(roc_perf_randfor@x.values[[1]]),
                               length(roc_perf_nn@x.values[[1]]),
                               length(roc_perf_opt@x.values[[1]])
                               )
                     )
                 )
)

ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("brown4", "purple", "cadetblue")) +
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", # (FPR)
       y = "True Positive Rate", # (TPR)
       color = "Model",
       linetype = "Model") +
  #theme_minimal() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") # Positions legend at the bottom
```






```{r echo=FALSE}
roc_data <- data.frame(
  FPR = c(
    unlist(roc_perf_randfor@x.values), # Random Forest
    unlist(roc_perf_lasso@x.values), # LASSO Logistic Regression
    unlist(roc_perf_opt@x.values) # Neural Net Optimized
    ),
  TPR = c(
    unlist(roc_perf_randfor@y.values), # Random Forest
    unlist(roc_perf_lasso@y.values), # LASSO Logistic Regression
    unlist(roc_perf_opt@y.values) # Neural Net Optimized
    ),
  Model = factor(rep(c("Random Forest", "LASSO", "Neural Net Optimized"), 
                     times = c(length(roc_perf_randfor@x.values[[1]]),
                               length(roc_perf_lasso@x.values[[1]]),
                               length(roc_perf_opt@x.values[[1]])
                               )
                     )
                 )
)

ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  scale_color_manual(values = c("green", "purple", "cadetblue")) +
  labs(title = "ROC Curve Comparison",
       x = "False Positive Rate", # (FPR)
       y = "True Positive Rate", # (TPR)
       color = "Model",
       linetype = "Model") +
  #theme_minimal() +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") # Positions legend at the bottom
```

**Answer:**

**Plotting the ROC Curve for all the methods, we see that our CV Decision Tree performs the worst, then Random Forest. After that, it's a matter of personal preference. Random Forest does outperform CV Decision Tree throughout the ROC Curve. In my opinion Decision Tree and Random Forest would not be suitable for our problem since we already narrowed it down to two candidates. I still think LASSO Logistic Regression is the most optimal, followed by our regular Logistic Regression.**
